---
title: "IFPH InFORM Join"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
library(viridis)
# install.packages("reticulate")
```

```{r}
df <- read_csv("data/processed_grid.csv")
shp <- st_read("data/processed_grid.shp")
```

```{r}
names(shp)
head(shp)

df <- as_tibble(shp)

shp <- shp %>% mutate(plant_sum = rowSums(df[,4:10]))

names(df)
head(df)

df <- df %>%
    mutate(
        is_indigenous = is_ndgn,
        n_quercus = n_qurcs,
        n_xerophyllum = n_xrphy,
        n_populus = n_popls,
        fire_lst = fir_lst,
    ) %>%
    select(-c(is_ndgn, n_qurcs, n_xrphy, n_popls, fir_lst)) %>%
    relocate(geometry, .after=fire_lst)

df <- df %>%
    mutate(plant_sum = rowSums(df[,4:10]), .after=n_populus)


df %>%
    select(plant_sum, n_quercus, n_xerophyllum, n_thuja, n_populus, n_rubus, n_pinus, n_salix, n_fires) %>%
    arrange(desc(plant_sum))

df %>%
    mutate(is_indigenous = as_factor(is_indigenous)) %>%
    ggplot(aes(geometry=geometry, fill=plant_sum, color=is_indigenous)) +
    geom_sf(linewidth=0.1) +
    scale_fill_viridis_c(option="inferno") +
    scale_colour_discrete(palette=c("white", "black")) +
    theme_void()

df %>%
    mutate(is_indigenous = as_factor(is_indigenous)) %>%
    ggplot(aes(geometry=geometry, fill=plant_sum, color=is_indigenous)) +
    geom_sf(linewidth=0.1) +
    scale_fill_binned(breaks=c(0, 0.5, 10, 100), palette=c("red", "blue", "navy", "black")) +
    scale_colour_discrete(palette=c("white", "black")) +
    theme_void()

summary(df$plant_sum)

```

```{r}
# fixed path using sudo chown -R
# hopefully. this will never happen again!
hist_veg <- st_read("/Users/nikolefiden/Desktop/Historic_Vegetation_84513795379117550/Historic_Vegetation.shp")

```

```{r}
head(hist_veg)

length(unique(hist_veg$VEGNAME))
as_tibble(hist_veg) %>% 
    group_by(VEGNAME) %>% 
    summarize(n=n()) %>% 
    arrange(desc(n)) %>%
    print(n=84)
```

Our GBIF species
- Ponderosa pine
- Oregon white oak
- Beargrass
- California blackberry
- Scouler's willow
- Western redcedar
- Black cottonwood

Of interest for us: 

Species 

- Ponderosa pine
    - Ponderosa pine-lodgepole pine
    - Douglas fir-ponderosa pine
- Willows
- White oak woodland
    - White oak-black oak
- Black cottonwood riparian woodland
- Western redcedar

Other

- Open water
- Riparian hardwoods
- Riparian
- Mixed conifer
- Marsh/Wetland
- Bare rock

```{r}
states <- usmap::us_map()
or <- states %>% filter(abbr=="OR")
# transforming OR so the coordinate system matches
or <- st_transform(or, crs="WGS84")
head(or)

st_crs(shp)

ponderosa_plt <- hist_veg %>%
    filter(VEGNAME %in% c("Ponderosa pine", "Ponderosa pine-lodgepole pine","Douglas fir-ponderosa pine")) %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="darkgreen") +
    geom_sf(data=shp %>% filter(n_pinus>0), fill="green", alpha=0.8) +
    theme_minimal()

hist_veg %>%
    filter(VEGNAME=="Ponderosa pine") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="darkgreen") +
    geom_sf(data=shp %>% filter(n_pinus>0), aes(fill=n_pinus), alpha=0.8) +
    scale_fill_binned(breaks=c(0, 1, 10, 100)) +
    theme_minimal()

ponderosa_plt

ggsave("ponderosa_plt.jpeg", ponderosa_plt)

```


```{r}
as_tibble(hist_veg) %>%
    filter(str_detect(VEGNAME, "oak"))

white_oak_plt <- hist_veg %>%
    filter(VEGNAME %in% c("White oak-black oak", "White oak woodland")) %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="darkgreen") +
    geom_sf(data=shp %>% filter(n_qurcs>0), fill="green", alpha=0.7) +
    theme_minimal()

white_oak_plt

ggsave("white_oak_plt.jpeg", white_oak_plt)

```


```{r}
as_tibble(hist_veg) %>%
    filter(str_detect(VEGNAME, "redcedar"))

western_redcedar_plt <- hist_veg %>%
    filter(VEGNAME=="Western redcedar") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="darkgreen") +
    geom_sf(data=shp %>% filter(n_thuja>0), fill="green", alpha=0.7) +
    theme_minimal()

western_redcedar_plt

ggsave("western_redcedar_plt.jpeg", western_redcedar_plt)
```


```{r}
as_tibble(hist_veg) %>%
    filter(str_detect(VEGNAME, "cottonwood"))

cottonwood_plt <- hist_veg %>%
    filter(VEGNAME=="Black cottonwood riparian woodland") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="darkgreen") +
    geom_sf(data=shp %>% filter(n_popls>0), fill="green", alpha=0.7) +
    theme_minimal()

cottonwood_plt

ggsave("cottonwood_plt.jpeg", cottonwood_plt)

```

```{r}
willows_plt <- hist_veg %>%
    filter(VEGNAME=="Willows") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="darkgreen") +
    geom_sf(data=shp %>% filter(n_salix>0), fill="green", alpha=0.7) +
    theme_minimal()

willows_plt

ggsave("willows_plt.jpeg", willows_plt)
```

```{r}
as_tibble(hist_veg) %>%
    filter(str_detect(VEGNAME, "Sand"))

hist_veg %>%
    filter(VEGNAME=="Sand dunes") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(color="tan", fill="yellow") +
    theme_minimal()

```

```{r}
hist_veg %>%
    filter(VEGNAME=="Bare rock") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="gray") +
    theme_minimal()
```

```{r}
hist_veg %>%
    filter(VEGNAME=="Open water") %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="blue", color="navy") +
    theme_minimal()
```

```{r}
brush_plt <- hist_veg %>%
    filter(str_detect(VEGNAME, "brush")) %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(fill="orange2", color="orange4") +
    geom_sf(data=shp %>% filter(plant_sum>0), fill="green", alpha=0.7) +
    theme_minimal()

brush_plt

ggsave("brush_plt.jpeg", brush_plt)

```

```{r}
vegname_fills <- c("Open water" = "blue", "Bare rock" = "gray", "Brush" = "orange2")

vegname_cols <- c("Open water" = "navy", "Bare rock" = "darkgray", "Brush" = "orange4")

noplant_plt <- hist_veg %>%
    filter(VEGNAME %in% c("Open water", "Bare rock") | str_detect(VEGNAME, "brush")) %>%
    mutate(VEGNAME = ifelse(str_detect(VEGNAME, "brush"), "brush", VEGNAME)) %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(aes(fill=VEGNAME, color=VEGNAME)) +
    scale_fill_manual(values=vegname_fills) +
    scale_color_manual(values=vegname_cols) +
    theme_minimal()

noplant_plt
```

```{r}
vegname_fills <- c("Open water" = "blue", "Bare rock" = "gray", "Brush" = "orange2", "Ponderosa pine" = "darkgreen")

vegname_cols <- c("Open water" = "navy", "Bare rock" = "darkgray", "Brush" = "orange4", "Ponderosa pine" = "black")

hist_veg %>%
    filter(VEGNAME %in% c("Open water", "Bare rock", "Ponderosa pine", "Ponderosa pine-lodgepole pine","Douglas fir-ponderosa pine") | str_detect(VEGNAME, "brush")) %>%
    mutate(
        VEGNAME = ifelse(str_detect(VEGNAME, "brush"), "Brush", VEGNAME),
        VEGNAME = ifelse(str_detect(VEGNAME, "pine"), "Ponderosa pine", VEGNAME)
    ) %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf(data=or, aes(geometry=geom)) +
    geom_sf(aes(fill=VEGNAME, color=VEGNAME), linewidth=0.1) +
    geom_sf(data=shp %>% filter(plant_sum>0), fill="green", alpha=0.7) +
    scale_fill_manual(values=vegname_fills) +
    scale_color_manual(values=vegname_cols) +
    theme_minimal()

```