---
title: "Create grid!"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
```

# Creating grid cells for oregon

## OR

```{r}
# read in oregon shapefile
states <- usmap::us_map()
or <- states %>% filter(abbr=="OR")
# transforming OR so the coordinate system matches
or <- st_transform(or, crs="WGS84")

```

## Create grid

```{r}
# make grid that covers OR area (note: it's a full rectangle)
# cellsize is based on lat/lon degrees
or_grid_square <- st_make_grid(or, cellsize=0.05) 

# test visualization
plot(or_grid_square)
plot(st_geometry(or), add=TRUE)
plot(or_grid_square[or], col="#ff000088", add=TRUE)
```

```{r}
head(or_grid_square)
# limiting to grid cells that are actually in oregon (e.g. removing non-red cells from prev visualization)
or_grid = or_grid_square[or]

# confirm it is smaller now (fewer cells)
length(or_grid_square)
length(or_grid)
```

## add names/identifier to each grid cell

```{r}
# creating index, writing as tibble
index = seq(1, length(or_grid), 1)
index_tib = tibble(index)

# using cbind to add index to grid (this transforms it to data.frame)
or_grid_df <- cbind(or_grid, index_tib)
class(or_grid_df)

```

# Add landmark data

```{r}
# read in landmark data
landmark <- st_read("/Users/nikolefiden/Desktop/DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_poly_public_v202509.shp")
# limit to USA, remove z-dimension, tranform coordinate system, make valid
us_lm <- landmark %>% filter(iso_code == "USA")
us_lm <- st_zm(us_lm)
us_lm <- st_transform(us_lm, crs="WGS84")
us_lm <- st_make_valid(us_lm)
```

## finding OR-intersecting areas

```{r}
# sf_use_s2(FALSE) (not sure why I had this before? doesn't seem necessary but if something breaks, it was)

# sparse geometry binary predicate list of length 1, where the predicate was "intersects"
# e.g. indexes where geometry intersects with OR
lm_int = st_intersects(or, us_lm)
dim(lm_int)

# only landmark rows that were in the intersection list
or_lm = us_lm[unlist(lm_int),]
dim(or_lm)
# 11 total
```

```{r}
# list for each cell and tribal area, TRUE if they intersect
indigenous <- st_intersects(or_grid, or_lm, sparse=FALSE)
indigenous_tibble <- tibble(indigenous)

or_grid_df <- cbind(or_grid_df, indigenous_tibble)

# for each grid cell, convert all indigenous rows to 1 if TRUE/0 if FALSE
# create is_indigenous column with sum of indigenous rows, so each cell has 1 if indigenous land and 0 if not
or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("indigenous"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        is_indigenous = ifelse(rowSums(across(starts_with("indigenous"))) == 0, 0, 1)
    )

# remove interim rows
or_grid_df <- or_grid_df %>% select(-indigenous)

lengths(or_grid_df)

```

# Add IFPH and InFORM data

```{r}
# reading in combined IFPH and InFORM data created in ifph-inform-join
fire_shp <- st_read("data/fire_shp.shp")

```

## Adding fire count

```{r}
# fire tibble with whether each fire occurred in each grid cell
fire <- st_intersects(or_grid, fire_shp, sparse=FALSE)
fire_tibble <- tibble(fire)
head(fire_tibble)
dim(fire_tibble)
dim(fire_tibble$fire)

# combining fire_tibble and or_grid_df to get counts of fires
or_grid_df <- cbind(or_grid_df, fire_tibble)

# creating sum of fires across each row/grid cell
or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("fire"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_fires = rowSums(across(starts_with("fire")))
    )

# removing the interim fire columns
or_grid_df <- or_grid_df %>% select(-fire)

head(or_grid_df %>%
    arrange(desc(n_fires))
)
# max number of fires: 8
```

## Adding fire identifier

```{r}
# getting grid index as list
grid_index <- or_grid_df$index

# new tibble that doesn't have a matrix
fire_tibble2 = cbind(tibble(data.frame(fire)), grid_index)
names(fire_tibble2)

# for every grid index and fire, shows the unique fire identifier (unique_id) and whether that fire was in that grid cell (is_fire)
long_fire <- fire_tibble2 %>%
    pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") %>%
    mutate(fire_index = as.numeric(str_remove(fire_index, "X"))) %>%
    mutate(
        unique_id = fire_shp$UNQE_FI[fire_index],
        acres = fire_shp$GIS_ACR[fire_index],
        fire_management_n = fire_shp$fr_mgm_[fire_index],
        duration_hrs = fire_shp$fr_df__[fire_index]
    )

# confirming the order didn't shift around for some reason
(long_fire %>% pull(unique_id))[1:100] == (fire_shp %>% pull(UNQE_FI))[1:100]

# creating new limited id column that has unique identifier if fire is present in that square, otherwise NA
# same for acres, fire_management_n, duration_hrs
# then pivoting so the unique ID shows in grid index col where the fire is present
wide_fire <- long_fire %>%
    # mutating unique ID to "missing" if NA and adding the constructed fire_index for referencing
    mutate(unique_id = ifelse(is.na(unique_id), paste("MISSING", as.symbol(fire_index)), unique_id)) %>%
    mutate(
        limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
    ) %>%
    pivot_wider(names_from = grid_index, values_from = limited_uid)

head(wide_fire)

# creating new tibble with grid index
# empty columns for unique identifiers, acres, mgmt complexity, duration 
fire_lst_tibble <- tibble(grid_index)
fire_lst_tibble[,"fire_lst"]=""
fire_lst_tibble[,"max_acres"]=""
fire_lst_tibble[,"min_mgmt"]=""
fire_lst_tibble[,"max_duration"]=""
head(fire_lst_tibble,5)

# for every grid cell, pulling unique fire identifiers for fires that occurred in that cell, adding them to fire_lst_tibble
for (x in 1:dim(or_grid_df)[1]) {
    # pull list of NAs and fire identifiers for each col/grid cell
    full_lst <- pull(wide_fire, x+6) # change 6 if more cols added
    # list of indexes for present fires
    fire_idx <- which(!is.na(full_lst))
    # list of present fire identifiers
    lst <- full_lst[!is.na(full_lst)]
    # if no fires, NA or 0 as applicable
    if (length(lst) == 0) {
        fire_lst_tibble$fire_lst[x] = NA
        fire_lst_tibble$max_acres[x] = 0
        fire_lst_tibble$min_mgmt[x] = NA
        fire_lst_tibble$max_duration[x] = 0
    # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
    } else {
        fire_lst_tibble$fire_lst[x] = paste(unlist(lst), collapse=", ")
        fire_lst_tibble$max_acres[x] = max(wide_fire$acres[fire_idx])
        fire_lst_tibble$min_mgmt[x] = min(wide_fire$fire_management_n[fire_idx])
        fire_lst_tibble$max_duration[x] = max(wide_fire$duration_hrs[fire_idx])
    }
}

# checking everything accurate
ex8 <- pull(wide_fire, `8`)
ex8_idx <- which(!is.na(ex8))
max(wide_fire$duration_hrs[ex8_idx])
ex8 <- ex8[!is.na(ex8)]
ex8
head(fire_lst_tibble,8)

ex8 == fire_lst_tibble$fire_lst[8]
```

```{r}
# joining or_grid_df and fire_lst_tibble on the grid index
or_grid_tib <- tibble(left_join(or_grid_df, fire_lst_tibble, by=c("index"="grid_index")))

or_grid_tib
```

```{r}
# test visualization
library(viridis)

or_grid_tib %>%
    mutate(is_indigenous = as_factor(is_indigenous)) %>%
    ggplot(aes(geometry=geometry, fill=n_fires, color=is_indigenous)) +
    geom_sf(linewidth=0.1) +
    scale_fill_viridis_c(option="inferno") +
    scale_colour_discrete(palette=c("gray", "black")) +
    theme_void()

```

```{r}
or_grid_tib %>%
    group_by(is_indigenous) %>%
    summarize(n=n())
```

# Adding GBIF data

## Read in and make sf

```{r}
gbif <- read_csv("data/new_gbif_dataset.csv")
```

```{r}
gbif_sf <- st_as_sf(gbif, coords=c("decimalLongitude", "decimalLatitude"), crs="WGS84")
gbif_sf <- gbif_sf %>% select(-`...1`)
```

```{r}
str(gbif_sf)

gbif_sf <- gbif_sf %>%
    mutate(
        species = as_factor(species),
        eventDate = ifelse(str_starts(eventDate, "20"), ymd_hms(eventDate, truncated=3), mdy_hms(eventDate, truncated=3)),
        basisOfRecord = as_factor(basisOfRecord),
        license = as_factor(license)
    ) 

```

# Species counts

```{r}
unique(gbif_sf$species)

# datasets filtered for each species
qg_gbif <- gbif_sf %>%
    filter(species == "Quercus garryana")

xt_gbif <- gbif_sf %>%
    filter(species == "Xerophyllum tenax")

tp_gbif <- gbif_sf %>%
    filter(species == "Thuja plicata")

pt_gbif <- gbif_sf %>%
    filter(species == "Populus trichocarpa")

ru_gbif <- gbif_sf %>%
    filter(species == "Rubus ursinus")

pp_gbif <- gbif_sf %>%
    filter(species == "Pinus ponderosa")

ss_gbif <- gbif_sf %>%
    filter(species == "Salix scouleriana")

```

```{r}
# contains lists/tibbles for each species
qg_con <- st_contains(or_grid, qg_gbif, sparse=FALSE)
qg_tibble <- tibble(qg_con)

xt_con <- st_contains(or_grid, xt_gbif, sparse=FALSE)
xt_tibble <- tibble(xt_con)

tp_con <- st_contains(or_grid, tp_gbif, sparse=FALSE)
tp_tibble <- tibble(tp_con)

pt_con <- st_contains(or_grid, pt_gbif, sparse=FALSE)
pt_tibble <- tibble(pt_con)

ru_con <- st_contains(or_grid, ru_gbif, sparse=FALSE)
ru_tibble <- tibble(ru_con)

pp_con <- st_contains(or_grid, pp_gbif, sparse=FALSE)
pp_tibble <- tibble(pp_con)

ss_con <- st_contains(or_grid, ss_gbif, sparse=FALSE)
ss_tibble <- tibble(ss_con)

head(qg_tibble)
```

```{r}
# adding counts per cell to overall df
# quercus garryana
or_grid_df <- cbind(or_grid_df, qg_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("qg_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_quercus = rowSums(across(starts_with("qg_con")))
    )

or_grid_df <- or_grid_df %>% select(-qg_con)

# xerophyllum tenax
or_grid_df <- cbind(or_grid_df, xt_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("xt_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_xerophyllum = rowSums(across(starts_with("xt_con")))
    )

or_grid_df <- or_grid_df %>% select(-xt_con)

# thuja plicata
or_grid_df <- cbind(or_grid_df, tp_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("tp_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_thuja = rowSums(across(starts_with("tp_con")))
    )

or_grid_df <- or_grid_df %>% select(-tp_con)

# populus trichocarpa
or_grid_df <- cbind(or_grid_df, pt_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("pt_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_populus = rowSums(across(starts_with("pt_con")))
    )

or_grid_df <- or_grid_df %>% select(-pt_con)

# rubus ursinus
or_grid_df <- cbind(or_grid_df, ru_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("ru_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_rubus = rowSums(across(starts_with("ru_con")))
    )

or_grid_df <- or_grid_df %>% select(-ru_con)

# pinus ponderosa
or_grid_df <- cbind(or_grid_df, pp_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("pp_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_pinus = rowSums(across(starts_with("pp_con")))
    )

or_grid_df <- or_grid_df %>% select(-pp_con)

# salix scouleriana
or_grid_df <- cbind(or_grid_df, ss_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("ss_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_salix = rowSums(across(starts_with("ss_con")))
    )

or_grid_df <- or_grid_df %>% select(-ss_con)

names(or_grid_df)
```

```{r}
# adding plant sums
or_grid_df <- or_grid_df %>% 
    mutate(plant_sum = rowSums(or_grid_df[,5:11])) # if cols added change 5:11

```

# Historic Vegetation info

```{r}
hist_veg <- st_read("/Users/nikolefiden/Desktop/Historic_Vegetation_84513795379117550/Historic_Vegetation.shp")
hist_veg <- st_transform(hist_veg, crs="WGS84")
```

## adding historic vegetation cols of interest

```{r}
# datasets filtered to each species
hist_pinus <- hist_veg %>%
    filter(VEGNAME %in% c("Ponderosa pine", "Ponderosa pine-lodgepole pine","Douglas fir-ponderosa pine"))

hist_salix <- hist_veg %>%
    filter(VEGNAME=="Willows")

hist_quercus <- hist_veg %>%
    filter(VEGNAME %in% c("White oak woodland", "White oak-black oak"))

hist_populus <- hist_veg %>%
    filter(VEGNAME=="Black cottonwood riparian woodland")

hist_thuja <- hist_veg %>%
    filter(VEGNAME=="Western redcedar")

```

```{r}
# contains lists/tibbles for each species
hist_pinus_grid <- st_intersects(or_grid, hist_pinus, sparse=FALSE)
hist_pinus_tibble <- tibble(hist_pinus_grid)

hist_salix_grid <- st_intersects(or_grid, hist_salix, sparse=FALSE)
hist_salix_tibble <- tibble(hist_salix_grid)

hist_quercus_grid <- st_intersects(or_grid, hist_quercus, sparse=FALSE)
hist_quercus_tibble <- tibble(hist_quercus_grid)

hist_populus_grid <- st_intersects(or_grid, hist_populus, sparse=FALSE)
hist_populus_tibble <- tibble(hist_populus_grid)

hist_thuja_grid <- st_intersects(or_grid, hist_thuja, sparse=FALSE)
hist_thuja_tibble <- tibble(hist_thuja_grid)

```

```{r}
# adding presence by cell to overall df
# pinus
or_grid_df <- cbind(or_grid_df, hist_pinus_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("hist_pinus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_pinus = ifelse(rowSums(across(starts_with("hist_pinus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_pinus_grid)

# salix
or_grid_df <- cbind(or_grid_df, hist_salix_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("hist_salix_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_salix = ifelse(rowSums(across(starts_with("hist_salix_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_salix_grid)

# quercus
or_grid_df <- cbind(or_grid_df, hist_quercus_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("hist_quercus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_quercus = ifelse(rowSums(across(starts_with("hist_quercus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_quercus_grid)

# populus 
or_grid_df <- cbind(or_grid_df, hist_populus_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("hist_populus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_populus = ifelse(rowSums(across(starts_with("hist_populus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_populus_grid)

# thuja
or_grid_df <- cbind(or_grid_df, hist_thuja_tibble)

or_grid_df <- or_grid_df %>%
    mutate(across(starts_with("hist_thuja_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_thuja = ifelse(rowSums(across(starts_with("hist_thuja_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_thuja_grid) 

names(or_grid_df)
```

## summing everything (GBIF and historic)

```{r}
or_grid_df <- or_grid_df %>%
    mutate(
        sum_pinus = rowSums(across(ends_with("pinus"))),
        sum_salix = rowSums(across(ends_with("salix"))),
        sum_quercus = rowSums(across(ends_with("quercus"))),
        sum_populus = rowSums(across(ends_with("populus"))),
        sum_thuja = rowSums(across(ends_with("thuja"))),
    ) %>% 
    # THIS NEEDS TO BE RERUN AND UPDATED
    # WAS NOT INCLUDING xeryophyllum or rubus initially
    mutate(
        total_sum = rowSums(across(c(starts_with("sum"), n_xerophyllum, n_rubus)))
    )

names(or_grid_df)
```

## Diversity index with vegan package

```{r}
div_df <- or_grid_df %>%
    select(sum_pinus, sum_salix, sum_quercus, sum_populus, sum_thuja, n_xerophyllum, n_rubus)

library(vegan)
shannon_div_tib <- tibble(diversity(div_df, index="shannon"))

simpson_div_tib <- tibble(diversity(div_df, index="simpson"))

or_grid_div <- cbind(or_grid_df, shannon_div_tib)
or_grid_div <- or_grid_div %>% rename(shannon_div = `diversity(div_df, index = \"shannon\")`)

or_grid_div <- cbind(or_grid_df, simpson_div_tib)
or_grid_div <- or_grid_div %>% rename(simpson_div = `diversity(div_df, index = \"simpson\")`)

names(or_grid_div)

```

```{r}
div_sf <- st_as_sf(or_grid_div, crs="WGS84")
names(div_sf)

div_sf %>%
    ggplot(aes(geometry=geometry, fill=simpson_div, alpha=n_fires)) +
    geom_sf(linewidth=0.1) +
    scale_fill_binned(breaks=c(0, 0, 0.01, 0.1, 1), palette=c("red", "black", "blue", "lightblue")) +
    scale_alpha_binned(range=c(0.2, 1), breaks=c(0, 0.5)) +
    theme_void()
```

# finalizing join

```{r}
full_grid_tib <- tibble(left_join(or_grid_df, fire_lst_tibble, by=c("index"="grid_index")))
```

```{r}
names(full_grid_tib)
full_grid_shp <- st_as_sf(full_grid_tib, crs="WGS84")

write_csv(full_grid_tib, "data/processed_grid.csv")
st_write(full_grid_shp, "data/processed_grid.shp", append=FALSE)

names(full_grid_tib)
```