---
title: "Fire Severity Feature Engineering"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
library(viridis)

```

# Reading in data

Created in inform-cleaning / ifph-cleaning

```{r}
inform <- read_csv("../DATA-510/inform_clean.csv")
ifph <- st_read("data/ifph_clean/ifph_clean.shp")

```

# Fire Severity Features

- Acres burned
- Burn duration
- Fire management complexity

## Acres burned

Already exists with no missingness - GIS_ACRES for IFPH, Calculated.Acres for InFORM

```{r}
dim(inform %>%
    filter(is.na(Calculated.Acres)))

dim(ifph %>%
    filter(is.na(GIS_ACRES)))
```

The National Wildfire Coordinating Group (NWCG) has size classifications for wildfires ([link](https://www.nwcg.gov/node/1737420)). We may limit smaller fires (which may represent minor incidents and not actual wildfires), so adding a factor for this classification.

Class A - one-fourth acre or less;
Class B - more than one-fourth acre, but less than 10 acres;
Class C - 10 acres or more, but less than 100 acres;
Class D - 100 acres or more, but less than 300 acres;
Class E - 300 acres or more, but less than 1,000 acres;
Class F - 1,000 acres or more, but less than 5,000 acres;
Class G - 5,000 acres or more.

```{r}
inform %>%
    select(Calculated.Acres) %>%
    arrange(desc(Calculated.Acres))
# max acres ~1050000 (a lot)

inform <- inform %>%
    mutate(size_class = cut(
        Calculated.Acres, 
        breaks=c(-0.01, 0.25, 9.99, 99.99, 299.99, 999.99, 4999.99, 9000000),
        labels = c("Class A", "Class B", "Class C", "Class D", "Class E", "Class F", "Class G")
    ))

inform %>%
    group_by(size_class) %>%
    summarize(n=n())

ifph <- ifph %>%
    mutate(size_class = cut(
        GIS_ACRES, 
        breaks=c(-0.01, 0.25, 9.99, 99.99, 299.99, 999.99, 4999.99, 9000000),
        labels = c("Class A", "Class B", "Class C", "Class D", "Class E", "Class F", "Class G")
    )) 

# running as a tibble so it doesn't take as long
tibble(ifph) %>%
    group_by(size_class) %>%
    summarize(n=n())

```

## Fire management complexity

Already exists - Fire.Mgmt.Complexity

```{r}
inform %>%
    group_by(Fire.Mgmt.Complexity) %>%
    summarize(n=n())
# many missing

inform %>%
    group_by(Fire.Mgmt.Complexity, Incident.Type.Category) %>%
    summarize(n=n()) %>%
    arrange(desc(n))
# some mismatch between things listed as Prescribed Fire in mgmt complexity vs incident type category
# to account for this I'll keep these two columns but also create fire_mgmt_n which extracts the number

inform <- inform %>%
    mutate(fire_mgmt_n = as.numeric(str_extract(Fire.Mgmt.Complexity, "[1-5]")))

inform %>%
    group_by(fire_mgmt_n) %>%
    summarize(n=n()) %>%
    arrange(desc(n))
# interesting that there are more 1s than 2s. also SIGNIFICANTLY missing :/

```

## Burn Duration

```{r}
inform %>%
    group_by(is.na(Containment.Date.Time), is.na(Control.Date.Time), is.na(FireOutDateTime)) %>%
    summarize(n=n()) %>%
    arrange(desc(n)) %>%
    print(width=Inf)
# 48088 missing all three - these will not have a burn duration unfortunately

inform %>% 
    mutate(no_duration = (is.na(Containment.Date.Time) & is.na(Control.Date.Time) & is.na(FireOutDateTime))) %>%
    group_by(no_duration, size_class) %>%
    summarize(n=n())
# missingness doesn't seem to be associated with particular fire size :/ 

```

```{r}
# this will take a long time to run it's a lot of calculations

inform <- inform %>%
    rowwise() %>%
    # finding latest fire out date
    mutate(latest_out = max(c_across(c(Containment.Date.Time, Control.Date.Time, FireOutDateTime)))) %>%
    # duration between fire discovery and latest_out
    mutate(duration_hrs = ifelse(is.na(latest_out), NA, interval(Fire.Discovery.Date.Time, latest_out)/dhours(x=1))) 

inform %>%
    group_by(size_class) %>%
    summarize(
        mean_duration = mean(duration_hrs, na.rm=TRUE),
        n=n(),
        pct_na = sum(is.na(duration_hrs))*100 / n
    )

```

```{r}
fire_shp <- ifph
object.size(fire_shp)
st_write(fire_shp, "data/fire_shp.shp")

fire_df <- inform
object.size(fire_df)
write_csv(fire_df, "../DATA-510/fire_df.csv")
```