---
title: "Joining Historic Vegetation"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
library(viridis)
```

```{r}
shp <- st_read("data/processed_grid.shp")
hist_veg <- st_read("/Users/nikolefiden/Desktop/Historic_Vegetation_84513795379117550/Historic_Vegetation.shp")
```

```{r}
df <- as_tibble(shp)

shp <- shp %>% mutate(plant_sum = rowSums(df[,4:10]))

shp <- st_transform(shp, "WGS84")

df <- df %>%
    mutate(
        is_indigenous = is_ndgn,
        n_quercus = n_qurcs,
        n_xerophyllum = n_xrphy,
        n_populus = n_popls,
        fire_lst = fir_lst,
    ) %>%
    select(-c(is_ndgn, n_qurcs, n_xrphy, n_popls, fir_lst)) %>%
    relocate(geometry, .after=fire_lst)

df <- df %>%
    mutate(plant_sum = rowSums(df[,4:10]), .after=n_populus)

hist_veg <- st_transform(hist_veg, "WGS84")

```

## Descriptions of mappings

Our GBIF species
- Ponderosa pine (n_pinus)
- Oregon white oak (n_qurcs)
- Beargrass (n_xrphy)
- California blackberry (n_rubus)
- Scouler's willow (n_salix)
- Western redcedar (n_thuja)
- Black cottonwood (n_popls)

Vegetation names of interest:
- Ponderosa pine
    - Ponderosa pine-lodgepole pine
    - Douglas fir-ponderosa pine
- Willows
- White oak woodland
    - White oak-black oak
- Black cottonwood riparian woodland
- Western redcedar

n_pinus <- Ponderosa pine, Ponderosa pine-lodgepole pine, Douglas fir-ponderosa pine
n_salix <- Willows
n_qurcs <- White oak woodland, White oak-black oak
n_popls <- Black cottonwood riparian woodland
n_thuja <- Western redcedar

```{r}
names(shp)
names(hist_veg)

hist_veg %>%
    filter(VEGNAME %in% c("Ponderosa pine", "Ponderosa pine-lodgepole pine","Douglas fir-ponderosa pine")) %>%
    summarize(n=n())
# 1430 shapes in here!

hist_pinus <- hist_veg %>%
    filter(VEGNAME %in% c("Ponderosa pine", "Ponderosa pine-lodgepole pine","Douglas fir-ponderosa pine"))

hist_pinus_grid <- st_intersects(shp, hist_pinus, sparse=FALSE)
hist_pinus_tibble <- tibble(hist_pinus_grid)

shp <- cbind(shp, hist_pinus_tibble)

shp <- shp %>%
    mutate(across(starts_with("hist_pinus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_pinus = ifelse(rowSums(across(starts_with("hist_pinus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_pinus_grid)

```

```{r}
hist_veg %>%
    filter(VEGNAME=="Willows") %>%
    summarize(n=n())
# 617 shapes

hist_salix <- hist_veg %>%
    filter(VEGNAME=="Willows")

hist_salix_grid <- st_intersects(shp, hist_salix, sparse=FALSE)
hist_salix_tibble <- tibble(hist_salix_grid)

shp <- cbind(shp, hist_salix_tibble)

shp <- shp %>%
    mutate(across(starts_with("hist_salix_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_salix = ifelse(rowSums(across(starts_with("hist_salix_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_salix_grid)
```

```{r}
hist_veg %>%
    filter(VEGNAME %in% c("White oak woodland", "White oak-black oak")) %>%
    summarize(n=n())
# 131 shapes

hist_qurcs <- hist_veg %>%
    filter(VEGNAME %in% c("White oak woodland", "White oak-black oak"))

hist_qurcs_grid <- st_intersects(shp, hist_qurcs, sparse=FALSE)
hist_qurcs_tibble <- tibble(hist_qurcs_grid)

shp <- cbind(shp, hist_qurcs_tibble)

shp <- shp %>%
    mutate(across(starts_with("hist_qurcs_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_qurcs = ifelse(rowSums(across(starts_with("hist_qurcs_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_qurcs_grid)
```

```{r}
hist_veg %>%
    filter(VEGNAME=="Black cottonwood riparian woodland") %>%
    summarize(n=n())
# 75 shapes

hist_popls <- hist_veg %>%
    filter(VEGNAME=="Black cottonwood riparian woodland")

hist_popls_grid <- st_intersects(shp, hist_popls, sparse=FALSE)
hist_popls_tibble <- tibble(hist_popls_grid)

shp <- cbind(shp, hist_popls_tibble)

shp <- shp %>%
    mutate(across(starts_with("hist_popls_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_popls = ifelse(rowSums(across(starts_with("hist_popls_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_popls_grid)
```

```{r}
hist_veg %>%
    filter(VEGNAME=="Western redcedar") %>%
    summarize(n=n())
# 11 shapes

hist_thuja <- hist_veg %>%
    filter(VEGNAME=="Western redcedar")

hist_thuja_grid <- st_intersects(shp, hist_thuja, sparse=FALSE)
hist_thuja_tibble <- tibble(hist_thuja_grid)

shp <- cbind(shp, hist_thuja_tibble)

shp <- shp %>%
    mutate(across(starts_with("hist_thuja_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_thuja = ifelse(rowSums(across(starts_with("hist_thuja_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_thuja_grid) 
```

## summing everything!

```{r}
shp <- shp %>%
    mutate(
        sum_pinus = rowSums(across(ends_with("pinus"))),
        sum_salix = rowSums(across(ends_with("salix"))),
        sum_qurcs = rowSums(across(ends_with("qurcs"))),
        sum_popls = rowSums(across(ends_with("popls"))),
        sum_thuja = rowSums(across(ends_with("thuja"))),
        .before=geometry
    ) %>% 
    mutate(
        total_sum = rowSums(across(starts_with("sum"))),
        .before=geometry
    )

head(shp)
```


```{r}
shp %>%
    ggplot(aes(geometry=geometry, fill=total_sum)) +
    geom_sf(linewidth=0.1) +
    scale_fill_binned(breaks=c(0, 0.5, 10, 100), palette=c("red", "blue", "navy", "black")) +
    theme_void()

shp %>%
    mutate(plants = ifelse(total_sum == 0, 0, 1)) %>%
    ggplot(aes(geometry=geometry, alpha=plants, fill=n_fires)) +
    geom_sf(linewidth=0.1) +
    scale_alpha(range=c(0.7, 1)) +
    #scale_alpha_binned(breaks=c(0, 0.9), range=c(0.6, 1.2)) +
    scale_fill_viridis(option="rocket") +
    theme_void()
```