---
title: "IFPH InFORM Join"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

# IFPH InFORM join

## Ingestion from download

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
```

```{r}
ifph <- st_read("/Users/nikolefiden/Desktop/DATA-510/InterAgencyFirePerimeterHistory_All_Years_View/InteragencyFirePerimeterHistory.shp")

inform <- read_csv("/Users/nikolefiden/Desktop/DATA-510/InFORM_FireOccurrence_Public.csv")
```

## Cleaning up

```{r}
# replacing spaces in col names with "." 
inform <- inform %>% dplyr::rename_all(~ make.names(.))
```

### Limiting to OR

Getting just OR from `usmap` to filter fires/native land to our area of interest with `st_intersects`. 

```{r}
states <- usmap::us_map()
or <- states %>% filter(abbr=="OR")
# transforming OR so the coordinate system matches
or <- st_transform(or, crs=st_crs(ifph))

or %>% 
    ggplot() +
    geom_sf() + 
    theme_minimal()
```

```{r}
lengths(ifph)

# creating intersection list for all polygons that intersect with OR border
# then selecting only those polygons
ifph_int = st_intersects(or, ifph)
dim(ifph_int)

or_ifph = ifph[unlist(ifph_int),]
dim(or_ifph)

or_ifph %>% 
    ggplot() + geom_sf() + theme_minimal()

lengths(or_ifph)
```

### Creating dataframe

```{r}
ifph_df <- as.data.frame(or_ifph)

```

### IFPH Cleaning

```{r}
# limiting FIRE_YEAR_ to 2015-2025 (inclusive)
ifph_df <- ifph_df %>%
    filter(FIRE_YEAR_ >= 2015, FIRE_YEAR_ <= 2025) 

# converting INCIDENT = UNNAMED, UNKNOWN to NA
ifph_df %>%
    mutate(INCIDENT = toupper(INCIDENT)) %>%
    group_by(INCIDENT) %>%
    summarize(n = n()) %>%
    arrange(desc(n))

ifph_df <- ifph_df %>%
    mutate(INCIDENT = toupper(INCIDENT)) %>%
    mutate(INCIDENT = ifelse(INCIDENT %in% c("UNNAMED", "UNKNOWN"), NA, INCIDENT)) 

ifph_df %>%
    mutate(INCIDENT = toupper(INCIDENT)) %>%
    group_by(INCIDENT) %>%
    summarize(n = n()) %>%
    arrange(desc(n))

# converting DATE_CUR to actual datetime 
# not sure what the timezone is. but it probably doesn't matter for our purposes
ifph_df <- ifph_df %>%
    mutate(DATE_CUR = lubridate::ymd_hms(DATE_CUR))

# shows an error for the NAs that they "failed to parse" - it's fine

# making FEATURE_CA into a factor
ifph_df <- ifph_df %>%
    mutate(FEATURE_CA = as_factor(FEATURE_CA)) 

# cleaning MAP_METHOD and turning into factor
print(n=30, ifph_df %>%
    mutate(MAP_METHOD = toupper(MAP_METHOD)) %>%
    group_by(MAP_METHOD) %>%
    summarize(n = n()) %>%
    arrange(desc(n)))

ifph_df <- ifph_df %>%
    mutate(MAP_METHOD = ifelse(MAP_METHOD %in% c("Mixed Method", "Mixed Collection Methods"), "Mixed Methods", MAP_METHOD),
        MAP_METHOD = ifelse(MAP_METHOD == "Digitized Other", "Digitized-Other", MAP_METHOD)) %>%
    mutate(MAP_METHOD = as_factor(MAP_METHOD))

# there are others that could be grouped (or are just the same) but for our purposes I'm not going to 

# making SOURCE into a factor
ifph_df <- ifph_df %>%
    mutate(SOURCE = as_factor(SOURCE)) 

# making AGENCY into a factor
ifph_df <- ifph_df %>%
    mutate(AGENCY = as_factor(AGENCY)) 

# removing FIRE_YEAR col
ifph_df <- ifph_df %>%
    select(-FIRE_YEAR)
```

## InFORM Cleaning

### Data type conversion

#### Numeric 

```{r}
# converting POO.Fips to numeric
inform <- inform %>%
    mutate(POO.Fips = as.numeric(POO.Fips)) 

```

#### Factor

```{r}
# changing ADS.Permission.State to a factor
inform <- inform %>%
    mutate(ADS.Permission.State = as_factor(ADS.Permission.State))

# Created.By.System -> factor
inform <- inform %>%
    mutate(Created.By.System = as_factor(Created.By.System))

# converting Dispatch.Center.ID to factor
inform <- inform %>% 
    mutate(Dispatch.Center.ID = as_factor(Dispatch.Center.ID))

# converting Fire.Mgmt.Complexity to ordered factor
inform %>% 
    group_by(Fire.Mgmt.Complexity) %>%
    summarize(n = n())

inform <- inform %>%
    mutate(Fire.Mgmt.Complexity = factor(Fire.Mgmt.Complexity, 
        levels=c(
            "Type 1 Incident",
            "Type 1 Prescribed Fire",
            "Type 2 Incident",
            "Type 3 Incident", 
            "Type 3 Prescribed Fire",
            "Type 4 Incident",
            "Type 5 Incident",
            "Complex Incident"
        ), ordered=TRUE))

# converting FS.Job.Code to factor
inform <- inform %>%
    mutate(FS.Job.Code = as_factor(FS.Job.Code))

# converting FS.Override.Code to factor
inform <- inform %>%
    mutate(FS.Override.Code = as_factor(FS.Override.Code))

# converting GACC to factor
inform <- inform %>%
    mutate(GACC = as_factor(GACC))

# converting Incident.Type.Category to factor
inform <- inform %>%
    mutate(Incident.Type.Category = as_factor(Incident.Type.Category))

# converting POO.Jurisdictional.Agency to factor
inform <- inform %>%
    mutate(POO.Jurisdictional.Agency = as_factor(POO.Jurisdictional.Agency))

# converting POOLandownerCategory to factor
inform <- inform %>%
    mutate(POOLandownerCategory = as_factor(POOLandownerCategory))

# converting POOLandownerKind to factor
inform <- inform %>%
    mutate(POOLandownerKind = as_factor(POOLandownerKind))

# converting POO.Protecting.Agency to factor
inform <- inform %>%
    mutate(POO.Protecting.Agency = as_factor(POO.Protecting.Agency))

# converting POO.State to factor
inform <- inform %>%
    mutate(POO.State = as_factor(POO.State))

# converting Predominant.Fuel.Group to factor
inform <- inform %>%
    mutate(Predominant.Fuel.Group = as_factor(Predominant.Fuel.Group))

# converting ModifiedBySystem to factor
inform <- inform %>%
    mutate(ModifiedBySystem = as_factor(ModifiedBySystem))

# converting Status to factor
inform <- inform %>%
    mutate(Status = as_factor(Status))
```

#### Logical

```{r}
# convert Is.Fire.Cause.Investigated to bool
inform <- inform %>% 
    mutate(Is.Fire.Cause.Investigated = as.logical(Is.Fire.Cause.Investigated))

# convert Is.FS.Assisted to bool
inform <- inform %>% 
    mutate(Is.FS.Assisted = as.logical(Is.FS.Assisted))

# convert Is.Trespass to bool
inform <- inform %>% 
    mutate(Is.Trespass = as.logical(Is.Trespass))

# convert IsCauseProhibited to bool
inform <- inform %>% 
    mutate(IsCauseProhibited = as.logical(IsCauseProhibited))
```

#### Datetime

```{r}
# converting Containment.Date.Time to a date
inform <- inform %>%
    mutate(Containment.Date.Time = lubridate::mdy_hms(Containment.Date.Time))

# converting Control.Date.Time to a date
inform <- inform %>%
    mutate(Control.Date.Time = lubridate::mdy_hms(Control.Date.Time))

# converting Created.On.Date.Time to datetime
inform <- inform %>%
    mutate(Created.On.Date.Time = as_datetime(Created.On.Date.Time/1000)) 

# converting Final.Fire.Report.Approved.Date to datetime
inform <- inform %>%
    mutate(Final.Fire.Report.Approved.Date = mdy_hms(Final.Fire.Report.Approved.Date))

# converting Fire.Discovery.Date.Time to datetime
inform <- inform %>%
    mutate(Fire.Discovery.Date.Time = mdy_hms(Fire.Discovery.Date.Time))

# converting FireOutDateTime to datetime
inform <- inform %>%
    mutate(FireOutDateTime = mdy_hms(FireOutDateTime))

# converting Initial.Response.Date.Time to datetime
inform <- inform %>%
    mutate(Initial.Response.Date.Time = mdy_hms(Initial.Response.Date.Time))

# converting Modified.On.Date.Time to datetime
inform <- inform %>%
    mutate(Modified.On.Date.Time = mdy_hms(Modified.On.Date.Time))

```

### Column removal, filtering

```{r}
# removing columns that are missing all values
inform <- inform %>% 
    select(-c(
        Discovery.Acres, 
        Estimated.Cost.To.Date, 
        Final.Acres, 
        Final.Fire.Report.Approved.By.Unit, 
        Fire.Behavior.General, 
        Incident.Short.Description, 
        Is.Reimbursable, 
        POO.Jurisdictional.Unit.Parent.Unit, 
        Percent.Contained
    ))
```

```{r}
# limiting Calendar.Year to just 2015-2026
# missing values for Calendar.Year are all present for Fire.Discovery.Date.Time - can impute

inform <- inform %>%
    mutate(
        # imputing values that are off by 100 - can assume 1922 -> 2022 and 2121 -> 2021
        Calendar.Year = ifelse(abs(year(Fire.Discovery.Date.Time) - Calendar.Year)==100, 
                        year(Fire.Discovery.Date.Time), 
                        Calendar.Year),
        # imputing missing values 
        Calendar.Year = ifelse(is.na(Calendar.Year), 
                        year(Fire.Discovery.Date.Time), 
                        Calendar.Year)
    ) 
    
inform %>%
    # removing all where Fire.Discovery.Date.Time and Calendar.Year are misaligned by more than one year
    filter(abs(year(Fire.Discovery.Date.Time) - Calendar.Year) <= 1) %>%
    filter(Calendar.Year >= 2015, Calendar.Year <= 2026)

inform %>%
    group_by(Calendar.Year) %>%
    summarize(n = n())
```

```{r}
# removing anomaly -1 acres burned fire
inform <- inform %>%
    filter(Calculated.Acres >= 0)
```

### Parsing/changing specific entry types to NA

```{r}
# 1970-01-01 is to be a placeholder, changing to NA
inform %>% 
    mutate(Containment.Date.Time = na_if(Containment.Date.Time, as_datetime("1970-01-01 00:00:00"))) %>%
    group_by(Containment.Date.Time) %>%
    summarize(n=n()) %>%
    arrange(Containment.Date.Time)

inform <- inform %>% 
    mutate(Containment.Date.Time = na_if(Containment.Date.Time, as_datetime("1970-01-01 00:00:00")))

```

```{r}
# cleaning Fire.Cause.Comments
inform %>%
    group_by(Fire.Cause.Comments) %>%
    summarize(n = n()) %>%
    arrange(Fire.Cause.Comments)

inform %>%
    mutate(Fire.Cause.Comments = str_remove(Fire.Cause.Comments, "\n")
    ) %>%
    mutate(Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("N/A", "NA", "", "*", ".", "-", ".."), NA, Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c(
                "UNKNOWN", 
                "UNDETERMINED", 
                "UNK", 
                "UN", 
                "UNKOWN", 
                "UNKNOWN CAUSE",
                "UK"
           ), "Unknown", Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("UNDER INVESTIGATION"), "Under investigation", Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("NOT INVESTIGATED"), "Not investigated", Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("NONE"), "None", Fire.Cause.Comments)
    ) %>%
    group_by(Fire.Cause.Comments) %>%
    summarize(n = n()) %>%
    arrange(desc(n)) %>%
    print(n=25)

inform <- inform %>% 
    mutate(Fire.Cause.Comments = str_remove(Fire.Cause.Comments, "\n")) %>%
    mutate(Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("N/A", "NA", "", "*", ".", "-", ".."), NA, Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c(
                "UNKNOWN", 
                "UNDETERMINED", 
                "UNK", 
                "UN", 
                "UNKOWN", 
                "UNKNOWN CAUSE",
                "UK"
           ), "Unknown", Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("UNDER INVESTIGATION"), "Under investigation", Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("NOT INVESTIGATED"), "Not investigated", Fire.Cause.Comments),
           Fire.Cause.Comments = ifelse(toupper(Fire.Cause.Comments) %in% c("NONE"), "None", Fire.Cause.Comments)
    )
```

# Joining them

```{r}
ifph_df %>%
    mutate(FORID = str_remove_all(FORID, "[{}]")) %>%
    mutate(
        unqe_match = ifelse(UNQE_FIRE_ %in% inform$UniqueFireIdentifier, T, F),
        forid_match = ifelse(FORID %in% inform$FORID, T, F),
        globalid_match = ifelse(GlobalID %in% inform$GlobalID, T, F)
    ) %>%
    group_by(unqe_match, forid_match, globalid_match)%>%
    summarize(n=n())

head(inform %>% select(UniqueFireIdentifier, FORID, GlobalID))
head(ifph_df %>% select(UNQE_FIRE_, FORID, GlobalID))

sum(ifph_df$UNQE_FIRE_ %in% inform$UniqueFireIdentifier)

```

```{r}
ifph_df %>%
    group_by(UNQE_FIRE_, GlobalID) %>%
    summarize(n=n()) %>%
    filter(n > 1) %>%
    arrange(desc(n))

inform %>%
    group_by(UniqueFireIdentifier, GlobalID) %>%
    summarize(n=n()) %>%
    filter(n > 1, !is.na(UniqueFireIdentifier)) %>%
    arrange(desc(n))

# everything with duplicated unique fire identifiers has a distinct global id for each entry. so i guess that makes sense. however the global IDs don't overlap across datasets
```

```{r}
ifph_df %>% 
    mutate(is_ifph = TRUE) %>%
    full_join(inform %>% mutate(is_inform = TRUE), by=c("UNQE_FIRE_" = "UniqueFireIdentifier")) %>%
    group_by(is_ifph, is_inform, Calendar.Year) %>%
    summarize(n=n()) %>%
    arrange(desc(Calendar.Year))


ifph_inform <- ifph_df %>% 
    mutate(is_ifph = TRUE) %>%
    left_join(inform %>% mutate(is_inform = TRUE), by=c("UNQE_FIRE_" = "UniqueFireIdentifier"))

ifph_inform %>% 
    group_by(is_inform, FIRE_YEAR_) %>%
    summarize(n=n()) %>%
    arrange(desc(n)) %>%
    print(n=20)

ifph_inform %>% 
    group_by(is_inform) %>%
    summarize(
        n=n(),
        quant_year = as_tibble_row(quantile(FIRE_YEAR_))
    ) %>%
    arrange(desc(n)) 
# this is a potential issue - the year spread is pretty distinct.
```

Not much to be done about it, but should be noted! 