---
title: "Data ingestion & validations"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

# Ingestion from download

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
```

```{r}
landmark <- st_read("/Users/nikolefiden/Desktop/DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_poly_public_v202509.shp")

ifph <- st_read("/Users/nikolefiden/Desktop/DATA-510/InterAgencyFirePerimeterHistory_All_Years_View/InteragencyFirePerimeterHistory.shp")

```

# Limiting to OR

Getting just OR from `usmap` to filter fires/native land to our area of interest with `st_intersects`. 

```{r}
states <- usmap::us_map()
or <- states %>% filter(abbr=="OR")
# transforming OR so the coordinate system matches
or <- st_transform(or, crs=st_crs(landmark))

or %>% 
    ggplot() +
    geom_sf() + 
    theme_minimal()
```

## LandMark for OR

```{r}
# issue with dimensions, removing z
landmark <- st_zm(landmark)
# limiting to just US first so it doesn't take as long to run
us_lm <- landmark %>% filter(iso_code == "USA")

lengths(us_lm)
head(us_lm)

# creating intersection list for all polygons that intersect with OR border
# then selecting only those polygons
lm_int = st_intersects(or, us_lm)
dim(lm_int)

or_lm = us_lm[unlist(lm_int),]
dim(or_lm)

or_lm %>% ggplot() + geom_sf()
lengths(or_lm)
```

## IFPH for OR
```{r}
lengths(ifph)

# creating intersection list for all polygons that intersect with OR border
# then selecting only those polygons
ifph_int = st_intersects(or, ifph)
dim(ifph_int)

or_ifph = ifph[unlist(ifph_int),]
dim(or_ifph)

# this is every recorded fire in OR over time - may take a second
or_ifph %>% ggplot() + geom_sf()
or_ifph %>% 
    filter(FIRE_YEAR_ >= 2015, FIRE_YEAR_ <= 2025) %>%
    ggplot() + geom_sf()
lengths(or_ifph)
```

# Overview

```{r}
# ifph overview
class(or_ifph)
head(or_ifph)
names(or_ifph)
str(or_ifph)
as.data.frame(or_ifph) %>% summary

# landmark overview
class(or_lm)
head(or_lm)
names(or_lm)
str(or_lm)
as.data.frame(or_lm) %>% summary
```

Lots of characters, so hard to see initially. There will be many conversions/transformations I think. IFPH years all look weird because they have historic data and unknowns marked as 9999. 

```{r}
# creating dataframes for each
ifph_df <- as.data.frame(or_ifph)
lm_df <- as.data.frame(or_lm)
```

# Missingness

## IFPH Missingness

```{r}
ifph_missing <- ifph_df %>%
    gather(key = "key", value = "val") %>%
    mutate(is_missing = is.na(val)) %>%
    group_by(key, is_missing) %>%
    summarize(num_missing = n()) %>%
    filter(is_missing==T) %>%
    select(-is_missing) %>%
    arrange(desc(num_missing))
    
```

```{r}
ifph_missing %>% 
    ggplot(aes(reorder(key, -num_missing), num_missing)) +
    geom_bar(stat="identity") +
    geom_hline(yintercept=pull(count(ifph_df))) +
    theme_minimal()
```


```{r}
ifph_missing_pct <- ifph_df %>%
    gather(key = "key", value = "val") %>%
    mutate(is_missing = is.na(val)) %>%
    group_by(key) %>%
    mutate(total = n()) %>%
    group_by(key, total, is_missing) %>%
    summarise(num_na = n()) %>%
    mutate(pct = num_na / total)

```


```{r}
levels <-
    (ifph_missing_pct %>% filter(is_missing == F) %>% arrange(pct))$key

ifph_missing_pct %>%
    ggplot(aes(key, pct, fill=is_missing)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=levels) +
    scale_y_continuous(labels=scales::percent) +
    scale_fill_manual(name = "", values = c('steelblue', 'tomato3'), labels = c("Present", "Missing")) +
    coord_flip() +
    theme_minimal()

```

Majority missing values for `IRWINID`, `COMMENTS`, `FORID`, `POO_RESP_I`. Significant missing values for `UNIT_ID`, `LOCAL_NUM`, and `UNQE_FIRE_`. I'll limit to 2015-2025 (our target timeframe) and test again. 

```{r}
ifph_missing_pct_current <- ifph_df %>%
    filter(FIRE_YEAR_ >= 2015, FIRE_YEAR_ <= 2025) %>%
    gather(key = "key", value = "val") %>%
    mutate(is_missing = is.na(val)) %>%
    group_by(key) %>%
    mutate(total = n()) %>%
    group_by(key, total, is_missing) %>%
    summarise(num_na = n()) %>%
    mutate(pct = num_na / total)

levels_current <-
    (ifph_missing_pct_current %>% filter(is_missing == F) %>% arrange(pct))$key

ifph_missing_pct_current %>%
    ggplot(aes(key, pct, fill=is_missing)) +
    geom_bar(stat="identity") +
    scale_x_discrete(limits=levels_current) +
    scale_y_continuous(labels=scales::percent) +
    scale_fill_manual(name = "", values = c('steelblue', 'tomato3'), labels = c("Present", "Missing")) +
    coord_flip() +
    theme_minimal()
```

Only majority missingness in comments, which is totally fine. Significant to moderate missingness for `POO_RESP_I`, `IRWINID`, `FORID`, `UNIT_ID`, `MAP_METHOD`, and `LOCAL_NUM`. Everything else looks pretty good within this timeframe. 

```{r}
dim(ifph_df %>%
    filter(FIRE_YEAR_ >= 2015, FIRE_YEAR_ <= 2025))

```

There are 1683 records within this time limit! That's a pretty decent showing. 

## LandMark Missingness

```{r}
lm_missing <- lm_df %>%
    gather(key = "key", value = "val") %>%
    mutate(is_missing = is.na(val)) %>%
    group_by(key, is_missing) %>%
    summarize(num_missing = n()) %>%
    filter(is_missing==T) %>%
    select(-is_missing) %>%
    arrange(desc(num_missing))

```

```{r}
lm_missing %>% 
    ggplot(aes(reorder(key, -num_missing), num_missing)) +
    geom_bar(stat="identity") +
    theme_minimal()
```

Not a particularly helpful graph, but it looks like there's minimal missing information. 

```{r}
lm_missing_pct <- lm_df %>%
    gather(key = "key", value = "val") %>%
    mutate(is_missing = is.na(val)) %>%
    group_by(key) %>%
    mutate(total = n()) %>%
    group_by(key, total, is_missing) %>%
    summarise(num_na = n()) %>%
    mutate(pct = num_na / total)

```


```{r}
lm_missing_pct %>%
    ggplot(aes(key, pct, fill=is_missing)) +
    geom_bar(stat="identity") +
    #scale_x_discrete(limits=levels) +
    scale_y_continuous(labels=scales::percent) +
    scale_fill_manual(name = "", values = c('steelblue', 'tomato3'), labels = c("Present", "Missing")) +
    coord_flip() +
    theme_minimal()

```

No partial missingness, there are only 11 records and all are missing `stat_note`, `stat_date`, `scale`, `pop_source`, `method`, `ethncty_1`, and `add_note`. The rest of the information is present for all records. For our purposes, this is totally fine. 

# Conclusion

In my opinion, these are in great shape for our purposes. The next major check will be to ensure the IFPH data plays nicely with the fire occurence records (but considering we're sourcing them from the same location, I have no reason to believe we'll face significant challenges there). LandMark only has 11 OR records, but given the nature of the information, I believe it is accurate and completely sufficient. 