---
title: "capstone test"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
```

# LandMark data

Initial test of LandMark polygon data:

```{r}
shp <- st_read("/Users/nikolefiden/Desktop/DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_poly_public_v202509.shp")
class(shp)
head(shp)
names(shp)
st_geometry(shp[1, ])[[1]][[1]]
plot(st_geometry(shp[1, ]))
```

All looks good, example shape plots appropriately.

```{r}
# checking dtype for each col for docs
str(shp)
```

Limiting the data to the USA, removing some Alaska Native Allotment, and plotting again (takes a lot longer to run). Looks good, plotted appropriately. 

```{r}
us_shp <- shp %>% filter(iso_code == "USA")
head(us_shp)
head(us_shp %>% filter(category != "Alaska Native Allotment"), 3)

unique(us_shp$category)

st_zm(us_shp) %>% filter(!st_is_empty(geometry), category != "Alaska Native Allotment") %>% ggplot() + geom_sf()


head(us_shp %>% filter(!st_is_empty(geometry), category != "Alaska Native Allotment"))
```

Plotting with the sf `plot` function instead of ggplot - also works. 

```{r}
plot(st_geometry(us_shp))
```

## LandMark Point Data

LandMark also has point data, which will be far less relevant to us, but can confirm it works as expected. 

```{r}
shppt <- st_read("/Users/nikolefiden/Desktop/DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_pt_public_v202509.shp")
class(shppt)
head(shppt)
st_geometry(shppt[1, ])[[1]][[1]]
plot(st_geometry(shppt[1, ]))
```

```{r}
us_shppt <- shppt %>% filter(iso_code == "USA")
head(us_shppt)
unique(us_shppt$category)
#ggplot(data=us_shp) + geom_sf()

us_shppt %>% filter(!st_is_empty(geometry)) %>% ggplot() + geom_sf()
```

# Interagency Fire Perimeter History

Moving on to IFPH data, which covers many years where the LM data is relatively static. 

```{r}
# i don't know why I called it IAFP instead of IFPH. don't worry about it
iafp <- st_read("/Users/nikolefiden/Desktop/DATA-510/InterAgencyFirePerimeterHistory_All_Years_View/InteragencyFirePerimeterHistory.shp")
class(iafp)
head(iafp)
st_geometry(iafp[1, ])[[1]][[1]]
plot(st_geometry(iafp[1, ]))
names(iafp)
```

Cols include a lot of things! Explained in the IFPH documentation. 

```{r}
lengths(iafp %>% filter(FIRE_YEAR_ <= 1900))
lengths(iafp %>% filter(FIRE_YEAR_ >= 2050))

```

2494 fires are listed at before 1900 - some are historical record, some were unknown so marked with an indicator. 268 fires are listed as later than 2050. 

```{r}
unique(iafp$SOURCE)

# Testing with FIRE_YEAR_ == 2022
iafp %>% filter(FIRE_YEAR_ == 2022) %>% ggplot() + geom_sf()
```

All looks good, that's the US!

# st_contains

Getting just OR from `usmap` to filter fires/native land to our area of interest with `st_contains`. 

```{r}
states<-usmap::us_map()
or = states %>% filter(abbr=="OR")
or %>% 
    ggplot()+
    geom_sf()

```

```{r}
# Filtering fires to 2020 for testing purposes (otherwise it is a lot of data)
iafp2020 <- iafp %>% filter(FIRE_YEAR_ == 2020)

or = st_transform(or, crs=st_crs(iafp2020))

con = st_contains(or, iafp2020)
con_sparse = st_contains(or, iafp2020, sparse=F)

iafp2020_or = iafp2020[unlist(con),]
head(iafp2020_or)

iafp2020_or %>% ggplot() + geom_sf()

```

As far as I can tell, that's the fires in Oregon in 2020!

```{r}
or = st_transform(or, crs=st_crs(iafp))

con = st_contains(or, iafp)
con_sparse = st_contains(or, iafp, sparse=F)

iafp_or = iafp[unlist(con),]

iafp_or <- iafp_or %>% filter(FIRE_YEAR_ >= 2000, FIRE_YEAR_ <= 2030)

lengths(iafp_or)

drop_na(as.data.frame(iafp_or$FIRE_YEAR_)) %>% summary

```

There are many fires! A lot of them are between 1981 and 2011, which is less relevant for our purposes, but that's ok. Limited to between 2000-2030 (max is 2024 anyway), and there are 4139. 

```{r}
intersects = st_intersects(or, iafp)

iafp_or_int = iafp[unlist(intersects),]

iafp_or_int <- iafp_or_int %>% filter(FIRE_YEAR_ >= 2000, FIRE_YEAR_ <= 2030)

iafp_or_int %>% filter(FIRE_YEAR_ == 2020) %>% ggplot() + geom_sf()

lengths(iafp_or_int)

drop_na(as.data.frame(iafp_or$FIRE_YEAR_)) %>% summary

```

This version includes slightly more fires, as it includes those that intersect but are not entirely contained within Oregon's border. We'll have to decide on our operational definitions for what cells contain fire/Indigenous land. 

