---
title: "capstone test"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
```

# LandMark data

Initial test of LandMark polygon data:

```{r}
shp <- st_read("/Users/nikolefiden/Desktop/DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_poly_public_v202509.shp")
class(shp)
head(shp)
names(shp)
st_geometry(shp[1, ])[[1]][[1]]
plot(st_geometry(shp[1, ]))
```

All looks good, example shape plots appropriately.

```{r}
# checking dtype for each col for docs
str(shp)
```

Limiting the data to the USA, removing some Alaska Native Allotment, and plotting again (takes a lot longer to run). Looks good, plotted appropriately. 

```{r}
us_shp <- shp %>% filter(iso_code == "USA")
head(us_shp)
head(us_shp %>% filter(category != "Alaska Native Allotment"), 3)

unique(us_shp$category)

st_zm(us_shp) %>% filter(!st_is_empty(geometry), category != "Alaska Native Allotment") %>% ggplot() + geom_sf()


head(us_shp %>% filter(!st_is_empty(geometry), category != "Alaska Native Allotment"))
```

Plotting with the sf native function instead of ggplot - also works. 

```{r}
plot(st_geometry(us_shp))
```

## LandMark Point Data

LandMark also has point data, which will be far less relevant to us, but can confirm it works as expected. 

```{r}
shppt <- st_read("/Users/nikolefiden/Desktop/DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_pt_public_v202509.shp")
class(shppt)
head(shppt)
st_geometry(shppt[1, ])[[1]][[1]]
plot(st_geometry(shppt[1, ]))
```

```{r}
us_shppt <- shppt %>% filter(iso_code == "USA")
head(us_shppt)
unique(us_shppt$category)
#ggplot(data=us_shp) + geom_sf()

us_shppt %>% filter(!st_is_empty(geometry)) %>% ggplot() + geom_sf()
```

# Interagency Fire Perimeter History

Moving on to IFPH data, which covers many years where the LM data is relatively static. 

```{r}
iafp <- st_read("/Users/nikolefiden/Desktop/DATA-510/InterAgencyFirePerimeterHistory_All_Years_View/InteragencyFirePerimeterHistory.shp")
class(iafp)
head(iafp)
st_geometry(iafp[1, ])[[1]][[1]]
plot(st_geometry(iafp[1, ]))
names(iafp)
```

Cols include a lot of things! Explained in the IAFP documentation. 

```{r}
iafp_hist <- iafp %>% filter(FIRE_YEAR_ <= 1900)
class(iafp_hist)

lengths(iafp_hist)

```

```{r}

unique(iafp$SOURCE)

iafp %>% filter(FIRE_YEAR_ == 2022) %>% ggplot() + geom_sf()

```

## st_contains

```{r}
states<-usmap::us_map()
or = states %>% filter(abbr=="OR")
or %>% 
    ggplot()+
    geom_sf()

```


```{r}
iafp2020 <- iafp %>% filter(FIRE_YEAR_ == 2020)

or = st_transform(or, crs=st_crs(iafp2020))

con = st_contains(or, iafp2020)
con_sparse = st_contains(or, iafp2020, sparse=F)

iafp2020_or = iafp2020[unlist(con),]
head(iafp2020_or)

iafp2020_or %>% ggplot() + geom_sf()


```
