---
title: "statistical tests!"
author: "Spencer Fiden"
date: "2026-08-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
library(viridis)
library(stats)
library(effectsize)
```

```{r}
gdf <- read_csv("./data/processed_grid/processed_grid.csv")
indigenous_cols <- c("Indigenous" = "salmon", "Non-Indigenous" = "slateblue3")

gdf <- gdf %>%
    mutate(is_ndgn = ifelse(is_indigenous==1, "Indigenous", "Non-Indigenous"))
```

# Indigenous land status 

```{r}
gdf %>%
    ggplot(aes(is_ndgn, max_acres, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ is_ndgn, data=gdf))
print(cohens_d(max_acres ~ is_ndgn, data=gdf))
```

```{r}
gdf %>% 
    filter(n_wildfires > 0) %>%
    ggplot(aes(is_ndgn, max_acres, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ is_ndgn, data=gdf %>% filter(n_wildfires > 0)))
```

Not significant at all.

```{r}
gdf %>% 
    ggplot(aes(is_ndgn, n_wildfires, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(n_wildfires ~ is_ndgn, data=gdf))
#print(t.test(n_wildfires ~ is_ndgn, data=gdf, alternative='less'))
print(cohens_d(n_wildfires ~ is_ndgn, data=gdf))

```

Some significance showing Indigenous land has higher counts of fires. 

```{r}
gdf %>% 
    filter(n_wildfires > 0) %>%
    ggplot(aes(is_ndgn, n_wildfires, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(n_wildfires ~ is_ndgn, data=gdf %>% filter(n_wildfires > 0)))
print(t.test(n_wildfires ~ is_ndgn, data=gdf %>% filter(n_wildfires > 0), alternative='less'))

print(cohens_d(n_wildfires ~ is_ndgn, data=gdf %>% filter(n_wildfires > 0)))

```

Some significance showing Indigenous land has higher counts of fires when limited to cells that have had fires.

```{r}
gdf %>% 
    ggplot(aes(is_ndgn, max_duration, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_duration ~ is_ndgn, data=gdf))
```

```{r}
gdf %>% 
    filter(n_wildfires > 0) %>%
    ggplot(aes(is_ndgn, max_duration, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_duration ~ is_ndgn, data=gdf %>% filter(n_wildfires > 0)))
```

Very little significance here. 

```{r}
gdf %>% 
    ggplot(aes(is_ndgn, n_rx_fires, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(n_rx_fires ~ is_ndgn, data=gdf))
```

No significance here either! Wow. 

```{r}
gdf %>% 
    filter(n_rx_fires > 0) %>%
    ggplot(aes(is_ndgn, n_rx_fires, fill=is_ndgn)) +
    geom_boxplot() +
    scale_fill_manual(values=indigenous_cols) +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(rx_max_acres ~ is_ndgn, data=gdf %>% filter(n_rx_fires > 0)))
print(cohens_d(rx_max_acres ~ is_ndgn, data=gdf %>% filter(n_rx_fires > 0)))

```

Technically some significance here but come on. Let's be real. 

# Previous wildfire/prescribed burn

```{r}
tgdf <- read_csv("./data/processed_grid/time_grid.csv")

tgdf <- tgdf %>%
    mutate(is_ndgn = ifelse(is_ndgn==1, "Indigenous", "Non-Indigenous"))
```

```{r}
# tgdf2 <- tgdf %>%
#     rowwise() %>%
#     mutate(
#         capped_acres = min(6178, max_acres)
#     ) 

tgdf <- tgdf %>%
    mutate(
        prev_rx_csum = rx_csum - n_rx,
        prev_rx_rollsum = rx_rollsum - n_rx,
        prev_fire_csum = fire_csum - n_fires,
        prev_fire_rollsum = fire_rollsum - n_fires,
        prev_fire_rollsum_5 = fire_rollsum_5 - n_fires
    )
```

## Rx csum

```{r}
tgdf %>% 
    ggplot(aes(as.factor(rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(rx_csum > 0), data=tgdf))
```

No significance. 

```{r}
tgdf %>% 
    filter(n_fires > 0) %>%
    ggplot(aes(as.factor(rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(rx_csum > 0), data=tgdf %>% filter(n_fires > 0)))
print(cohens_d(max_acres ~ as.factor(rx_csum > 0), data=tgdf %>% filter(n_fires > 0)))

```

Some significance for cumulative sum of prescribed burn when limited to fires existing! p value of 0.04.

Limiting to 2020-2025, as there aren't any prescribed burns before that. 

```{r}
tgdf %>% 
    filter(year >= 2020) %>%
    ggplot(aes(as.factor(rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(rx_csum > 0), data=tgdf %>% filter(year >= 2020)))

```

No significance at all! Wow. 

```{r}
tgdf %>% 
    filter(year >= 2020) %>%
    filter(n_fires > 0) %>%
    ggplot(aes(as.factor(rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(rx_csum > 0), data=tgdf %>% filter(n_fires > 0, year >= 2020)))
print(cohens_d(max_acres ~ as.factor(rx_csum > 0), data=tgdf %>% filter(n_fires > 0, year >= 2020)))

```

Quite significant - p-value of 0.002. 

## Previous rx csum

```{r}
tgdf %>% 
    ggplot(aes(as.factor(prev_rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_rx_csum > 0), data=tgdf))
```

Not significant.

```{r}
tgdf %>% 
    filter(n_fires > 0) %>%
    ggplot(aes(as.factor(prev_rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_rx_csum > 0), data=tgdf %>% filter(n_fires > 0)))
```

Almost significant - p-value of 0.057.

Limiting to 2020-2025, as there aren't any prescribed burns before that. 

```{r}
tgdf %>% 
    filter(year >= 2020) %>%
    ggplot(aes(as.factor(prev_rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_rx_csum > 0), data=tgdf %>% filter(year >= 2020)))
```

Not significant at all. 

```{r}
tgdf %>% 
    filter(year >= 2020) %>%
    filter(n_fires > 0) %>%
    ggplot(aes(as.factor(prev_rx_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_rx_csum > 0), data=tgdf %>% filter(n_fires > 0, year >= 2020)))
print(cohens_d(max_acres ~ as.factor(prev_rx_csum > 0), data=tgdf %>% filter(n_fires > 0, year >= 2020)))

```

Quite significant! p-value of 0.006. Cohen's d of 0.21, 95% CI [0.0, 0.4]

## Previous fire csum

```{r}
tgdf %>% 
    ggplot(aes(as.factor(prev_fire_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_fire_csum > 0), data=tgdf))
print(cohens_d(max_acres ~ as.factor(prev_fire_csum > 0), data=tgdf))

```

Almost significant - p-value of 0.054. This one is in the opposite direction - cells that have had fires are more likely to have slightly larger fires in the future. Cohen's d is -0.01. 

```{r}
tgdf %>% 
    filter(n_fires > 0) %>%
    ggplot(aes(as.factor(prev_fire_csum > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_fire_csum > 0), data=tgdf %>% filter(n_fires > 0)))
print(cohens_d(max_acres ~ as.factor(prev_fire_csum > 0), data=tgdf %>% filter(n_fires > 0)))

```

Very significant - p-value < 2.2e-16. Cohen's d of 0.26, CI [0.21, 0.30].

## Previous fire rollsum_5

```{r}
tgdf %>% 
    ggplot(aes(as.factor(prev_fire_rollsum_5 > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_fire_rollsum_5 > 0), data=tgdf))
print(cohens_d(max_acres ~ as.factor(prev_fire_rollsum_5 > 0), data=tgdf))

```

Not significant.

```{r}
tgdf %>% 
    filter(n_fires > 0) %>%
    ggplot(aes(as.factor(prev_fire_rollsum_5 > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_fire_rollsum_5 > 0), data=tgdf %>% filter(n_fires > 0)))
print(cohens_d(max_acres ~ as.factor(prev_fire_rollsum_5 > 0), data=tgdf %>% filter(n_fires > 0)))

```

Very significant - p-value < 2.2e-16 again. 

Limiting to 2020 and on so the full 5 years are visible. 

```{r}
tgdf %>% 
    filter(n_fires > 0) %>%
    filter(year >= 2020) %>%
    ggplot(aes(as.factor(prev_fire_rollsum_5 > 0), max_acres)) +
    geom_boxplot() +
    scale_y_continuous(trans=scales::pseudo_log_trans(base = 10)) +
    theme_bw()

print(t.test(max_acres ~ as.factor(prev_fire_rollsum_5 > 0), data=tgdf %>% filter(n_fires > 0, year >= 2020)))
print(cohens_d(max_acres ~ as.factor(prev_fire_rollsum_5 > 0), data=tgdf %>% filter(n_fires > 0, year >= 2020)))

```

Similarly very significant. 

