---
title: "Time grid eda"
author: "Spencer Fiden"
date: "2026-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(ggplot2)
library(dplyr)
library(zoo)
library(ranger)
library(Metrics)
library(caret)
```

```{r}
tg <- st_read("../DATA-510/time_grid/time_grid.shp")
tgdf <- read_csv("data/processed_grid/time_grid.csv")
```

```{r}
names(tgdf)
# creating limited df with only factors we'll use for modeling
lim_tg <- tgdf %>% select(
  year, index,
  is_ndgn, max_acres_tribal,
  n_pinus, n_salix, n_qurcs, n_popls, n_thuja, n_xrphy, n_rubus, plnt_sm, 
  n_fires, max_acres, max_duration, min_mgmt,
  n_rx, max_acres_rx, max_duration_rx, min_mgmt_rx
)

lim_tg <- lim_tg %>%
  mutate(
    rx_rollsum = rollapply(n_rx, 2, sum, partial=TRUE, na.rm=TRUE),
    rx_csum = cumsum(n_rx)
  )

```

```{r}
lim_tg %>%
  mutate(had_fire = (n_fires > 0) | (n_rx > 0)) %>%
  group_by(had_fire) %>%
  summarize(
    n=n(),
    mean_rx_rollsum = mean(rx_rollsum),
    mean_fires = mean(n_fires),
    mean_rx = mean(n_rx)
  )

lim_tg %>%
  filter((n_fires > 0) | (n_rx > 0))

# WAIT. this is per cell, not over time. hold on. 
lim_tg <- lim_tg %>%
  group_by(index) %>%
  mutate(
    sum_fires = sum(n_fires),
    sum_rx = sum(n_rx)
  )

lim_tg %>%
  mutate(had_fire = (sum_fires > 0) | (sum_rx > 0)) %>%
  group_by(had_fire) %>%
  summarize(
    n=n(),
    mean_rx_rollsum = mean(rx_rollsum),
    avg_ann_fires = mean(n_fires),
    avg_ann_rx = mean(n_rx)
  ) 

lim_tg %>%
  mutate(
    had_fire = sum_fires > 0,
    had_rx = sum_rx > 0
  ) %>%
  group_by(had_fire, had_rx) %>%
  summarize(
    n=n(),
    mean_rx_rollsum = mean(rx_rollsum),
    avg_ann_fires = mean(n_fires),
    avg_ann_rx = mean(n_rx)
  ) 

# dataset which only includes cells that have ever had a fire (prescribed or wildfire)
lim_tg_fire <- lim_tg %>%
  filter((sum_fires > 0) | (sum_rx > 0))

```

## Linear Regression

### All cells (including those that have never had fire)

```{r}
lm_acres_plant <- lm(
  max_acres ~
    n_pinus + n_salix + n_qurcs + n_popls + n_thuja + n_xrphy + n_rubus + rx_rollsum,
  data = lim_tg
)

summary(lm_acres_plant)

errors_acres_plant <- lm_acres_plant$residuals
rmse_plant <- sqrt(mean(errors_acres_plant^2))
mae_plant <- mean(abs(errors_acres_plant))

print(paste("RMSE:", rmse_plant))
print(paste("MAE:", mae_plant))
```

RMSE: ~18473
MAE: ~3045
Best Pr: 0.0336 (rx_rollsum, t value 2.125)

```{r}
lm_acres <- lm(
  max_acres ~
    plnt_sm + rx_rollsum + is_ndgn,
  data = lim_tg
)

summary(lm_acres)

errors_acres <- lm_acres$residuals
rmse <- sqrt(mean(errors_acres^2))
mae <- mean(abs(errors_acres))

print(paste("RMSE:", rmse))
print(paste("MAE:", mae))
```

RMSE: ~18474
MAE: ~3042
Best Pr: 0.00833 (plnt_sm, t value -2.638)

```{r}
lm_acres2 <- lm(
  max_acres ~
    (plnt_sm + rx_csum + is_ndgn)^2,
  data = lim_tg
)

summary(lm_acres2)

errors_acres2 <- lm_acres2$residuals
rmse2 <- sqrt(mean(errors_acres2^2))
mae2 <- mean(abs(errors_acres2))

print(paste("RMSE:", rmse2))
print(paste("MAE:", mae2))

```

RMSE: ~18469
MAE: ~3042
Best Pr: 4.10e-12 (rx_csum, t value 6.935)
(Followed by rx_csum:is_ndgn, 2.15e-05, t value -4.248)

```{r}
lm_acres3 <- lm(
  max_acres ~
    (plnt_sm + rx_csum + is_ndgn)^3,
  data = lim_tg
)

summary(lm_acres3)

errors_acres3 <- lm_acres3$residuals
rmse3 <- sqrt(mean(errors_acres3^2))
mae3 <- mean(abs(errors_acres3))

print(paste("RMSE:", rmse3))
print(paste("MAE:", mae3))
```

RMSE: ~18469
MAE: ~3042
Best Pr: 3.88e-12 (rx_csum, t value 6.942)

```{r}
plot(lm_acres2)
```

### Fire cells

```{r}
lm_fire_plant <- lm(
  max_acres ~
    n_pinus + n_salix + n_qurcs + n_popls + n_thuja + n_xrphy + n_rubus + rx_rollsum,
  data = lim_tg_fire
)

summary(lm_acres_plant)

errors_fire_plant <- lm_fire_plant$residuals
rmse_fire_plant <- sqrt(mean(errors_fire_plant^2))
mae_fire_plant <- mean(abs(errors_fire_plant))

print(paste("RMSE:", rmse_fire_plant))
print(paste("MAE:", mae_fire_plant))
```

RMSE: ~25744
MAE: ~5880
Best Pr: 0.0336 (rx_rollsum, t value 2.125)

```{r}
lm_fire <- lm(
  max_acres ~
    plnt_sm + rx_rollsum + is_ndgn,
  data = lim_tg_fire
)

summary(lm_fire)

errors_fire <- lm_fire$residuals
rmse_fire <- sqrt(mean(errors_fire^2))
mae_fire <- mean(abs(errors_fire))

print(paste("RMSE:", rmse_fire))
print(paste("MAE:", mae_fire))
```

RMSE: ~25746
MAE: ~5874
Best Pr: 0.000216 (plnt_sm, t value -3.699)

```{r}
lm_fire2 <- lm(
  max_acres ~
    (plnt_sm + rx_csum + is_ndgn)^2,
  data = lim_tg_fire
)

summary(lm_fire2)

errors_fire2 <- lm_fire2$residuals
rmse_fire2 <- sqrt(mean(errors_fire2^2))
mae_fire2 <- mean(abs(errors_fire2))

print(paste("RMSE:", rmse_fire2))
print(paste("MAE:", mae_fire2))

```

RMSE: ~25733
MAE: ~5869
Best Pr: 5.85e-13 (rx_csum, r value 7.206)

```{r}
# using rx_rollsum instead of cumulative
lm_fire2_roll <- lm(
  max_acres ~
    (plnt_sm + rx_rollsum + is_ndgn)^2,
  data = lim_tg_fire
)

summary(lm_fire2_roll)

errors_fire2_roll <- lm_fire2_roll$residuals
rmse_fire2_roll <- sqrt(mean(errors_fire2_roll^2))
mae_fire2_roll <- mean(abs(errors_fire2_roll))

print(paste("RMSE:", rmse_fire2_roll))
print(paste("MAE:", mae_fire2_roll))

```

RMSE: ~25746
MAE: ~5874
Best Pr: 0.000391 (plnt_sm, r value -3.546)

```{r}
lm_fire3 <- lm(
  max_acres ~
    (plnt_sm + rx_csum + is_ndgn)^3,
  data = lim_tg_fire
)

summary(lm_fire3)

errors_fire3 <- lm_fire3$residuals
rmse_fire3 <- sqrt(mean(errors_fire3^2))
mae_fire3 <- mean(abs(errors_fire3))

print(paste("RMSE:", rmse_fire3))
print(paste("MAE:", mae_fire3))
```

RMSE: ~25733
MAE: ~5869
Best Pr: 4.95e-13 (rx_csum, r value 7.228)

```{r}
plot(lm_fire2)
```

## Random Forest

### All Cells

note: should also try with fire class as a classification problem instead of regression.

```{r}
set.seed(42)

repeat_cv <- trainControl(method="repeatedcv", number=3, repeats=2, verboseIter=TRUE)

train_index <- createDataPartition(
  lim_tg$max_acres,
  p = .7,
  list = FALSE
)

train <- lim_tg[train_index, ]
test <- lim_tg[-train_index, ]
```

```{r}
# random forest for max acres

rf_acres <- ranger(
  max_acres ~
    plnt_sm + rx_csum + is_ndgn,
  data = train,
  importance = "permutation",
  num.trees = 500
)

# predict
pred_rf <- predict(rf_acres, data = test)$predictions

# evaluate
rmse(test$max_acres, pred_rf) #17675.15

mae(test$max_acres, pred_rf) #2954.363

cor(test$max_acres, pred_rf)^2 #0.07223494

#variable importance
importance(rf_acres)

#cross validation
control <- trainControl(
  method = "cv",
  number = 10
)

rf_cv <- train(
  max_acres ~
    plnt_sm + rx_csum + is_ndgn,
  data = lim_tg,
  method = "ranger",
  trControl = repeat_cv,
  importance = "permutation"
)

rf_cv

# R squared at best is 0.095, with mtry 2 splitrule variance
# not good!
```


```{r}
repeat_cv_long <- trainControl(method="repeatedcv", number=5, repeats=3, verboseIter=TRUE)

# random forest for max acres

rf_acres2 <- ranger(
  max_acres ~
    plnt_sm + rx_csum + is_ndgn,
  data = train,
  importance = "permutation",
  num.trees = 250
)

# predict
pred_rf2 <- predict(rf_acres2, data = test)$predictions

# evaluate
rmse(test$max_acres, pred_rf2) #17675.15

mae(test$max_acres, pred_rf2) #2954.363

cor(test$max_acres, pred_rf2)^2 #0.07223494

#variable importance
importance(rf_acres2)

#cross validation
rf_cv2 <- train(
  max_acres ~
    plnt_sm + rx_csum + is_ndgn,
  data = lim_tg,
  method = "ranger",
  trControl = repeat_cv_long,
  importance = "permutation"
)

rf_cv2
```

### Fire Cells

```{r}
train_index_f <- createDataPartition(
  lim_tg_fire$max_acres,
  p = .7,
  list = FALSE
)

train_f <- lim_tg_fire[train_index_f, ]
test_f <- lim_tg_fire[-train_index_f, ]
```

```{r}
# random forest for max acres

rf_acres_f <- ranger(
  max_acres ~
    plnt_sm + rx_csum + is_ndgn,
  data = train_f,
  importance = "permutation",
  num.trees = 500
)

# predict
pred_rf_f <- predict(rf_acres_f, data = test_f)$predictions

# evaluate
rmse(test_f$max_acres, pred_rf_f) #25673.89

mae(test_f$max_acres, pred_rf_f) #5734.942

cor(test_f$max_acres, pred_rf_f)^2 #0.1216258

#variable importance
importance(rf_acres_f)

#cross validation
control <- trainControl(
  method = "cv",
  number = 10
)

rf_cv_f <- train(
  max_acres ~
    plnt_sm + rx_csum + is_ndgn,
  data = lim_tg_fire,
  method = "ranger",
  trControl = repeat_cv,
  importance = "permutation"
)

rf_cv_f

# R squared at best is 0.1351947, with mtry 2 splitrule variance
# RMSE 23975.77, MAE 5144.613
```

