---
title: "Create grid!"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

Creates a processed grid with aggregated information from all of our data sources. 

Reads from:
- LandMark: local system (file too large)
- IFPH/InFORM: `data/ifph_inform_multipolygon`
- InFORM: `data/ifph_inform_point`
- GBIF: `data/new_gbif_dataset.csv`
- Historic Vegetation: local system (file too large)

Writes to `data/processed_grid`. (Includes verbose version with all columns and regular version with selected & renamed columns).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
library(viridis)

```

# Creating grid cells for oregon

## OR

```{r}
# read in oregon shapefile
states <- usmap::us_map()
or <- states %>% filter(abbr=="OR")
```

## Create grid

```{r}
# make grid that covers OR area (note: it's a full rectangle)
# cellsize is 5km by 5km (using us_map original coordinate system)
or_grid_square <- st_make_grid(or, cellsize=5000) 

# test visualization
plot(or_grid_square)
plot(st_geometry(or), add=TRUE)
plot(or_grid_square[or], col="#ff000088", add=TRUE)
```

```{r}
head(or_grid_square)
# limiting to grid cells that are actually in oregon (e.g. removing non-red cells from prev visualization)
or_grid = or_grid_square[or]

# transforming OR so the coordinate system matches our other data
or_grid <- st_transform(or_grid, crs="WGS84")

# confirm it is smaller now (fewer cells)
length(or_grid_square)
length(or_grid)
```

## add names/identifier to each grid cell

```{r}
# creating index, writing as tibble
index = seq(1, length(or_grid), 1)
index_tib = tibble(index)

# using cbind to add index to grid (this transforms it to data.frame)
or_grid_df <- cbind(or_grid, index_tib)
class(or_grid_df)

```

# Add landmark data

```{r}
# read in landmark data
landmark <- st_read("../DATA-510/Indigenous_comm_lands_v202509/LandMark_IPLC_poly_public_v202509.shp")
# limit to USA, remove z-dimension, tranform coordinate system, make valid
us_lm <- landmark %>% filter(iso_code == "USA")
us_lm <- st_zm(us_lm)
us_lm <- st_transform(us_lm, crs="WGS84")
us_lm <- st_make_valid(us_lm)
```

## finding OR-intersecting areas

```{r}
# sf_use_s2(FALSE) (not sure why I had this before? doesn't seem necessary but if something breaks, it was)

# sparse geometry binary predicate list of length 1, where the predicate was "intersects"
# e.g. indexes where geometry intersects with OR
lm_int = st_intersects(st_transform(or, crs="WGS84"), us_lm)
dim(lm_int)

# only landmark rows that were in the intersection list
or_lm = us_lm[unlist(lm_int),]
dim(or_lm)
# 11 total
```

```{r}
# list for each cell and tribal area, TRUE if they intersect
indigenous <- st_intersects(or_grid, or_lm, sparse=FALSE)
indigenous_tibble <- tibble(indigenous)

or_grid_ind <- cbind(or_grid_df, indigenous_tibble)

# for each grid cell, convert all indigenous rows to 1 if TRUE/0 if FALSE
# create is_indigenous column with sum of indigenous rows, so each cell has 1 if indigenous land and 0 if not
or_grid_ind <- or_grid_ind %>%
    mutate(across(starts_with("indigenous"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        is_indigenous = ifelse(rowSums(across(starts_with("indigenous"))) == 0, 0, 1)
    )

# remove interim rows
or_grid_ind <- or_grid_ind %>% select(-indigenous)

lengths(or_grid_ind)

```

# Add IFPH and InFORM data

```{r}
# reading in combined IFPH and InFORM data created in ifph-inform-join
full_fire_mp <- st_read("data/ifph_inform_multipolygon/ifph_inform_multipolygon.shp")

# reading in point data from InFORM lat/lon data (also created in ifph-inform-join)
full_fire_pt <- st_read("data/ifph_inform_point/ifph_inform_point.shp")

# separating into RX and WF
# note that this removes fires that were present in IFPH but not InFORM (we would be missing duration/mgmt complexity info for these as well)
rx_fire_mp <- full_fire_mp %>%
    filter(Inc_T_C == "RX")

fire_mp <- full_fire_mp %>%
    filter(Inc_T_C != "RX")

rx_fire_pt <- full_fire_pt %>%
    filter(Inc_T_C == "RX")

fire_pt <- full_fire_pt %>%
    filter(Inc_T_C != "RX")

```

## Adding IFPH/InFORM wildfire data

### Adding fire count

```{r}
# fire tibble with whether each fire occurred in each grid cell
fire_mp_int <- st_intersects(or_grid, fire_mp, sparse=FALSE)
fire_mp_tibble <- tibble(fire_mp_int)
head(fire_mp_tibble)
dim(fire_mp_tibble)
dim(fire_mp_tibble$fire_mp_int)

# fire tibble for the point geometries
fire_pt_int <- st_contains(or_grid, fire_pt, sparse=FALSE)
fire_pt_tibble <- tibble(fire_pt_int)

# combining fire_tibble and or_grid_df to get counts of fires
or_grid_fire <- cbind(or_grid_ind, fire_mp_tibble)

# creating sum of fires across each row/grid cell
or_grid_fire <- or_grid_fire %>%
    mutate(across(starts_with("fire"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_fires_mp = rowSums(across(starts_with("fire")))
    )

# removing the interim fire columns
or_grid_fire <- or_grid_fire %>% select(-fire_mp_int)

or_grid_fire2 <- cbind(or_grid_fire, fire_pt_tibble)

or_grid_fire2 <- or_grid_fire2 %>%
    mutate(across(starts_with("fire"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_fires_pt = rowSums(across(starts_with("fire")))
    )

or_grid_fire2 <- or_grid_fire2 %>% select(-fire_pt_int)

names(or_grid_fire2)

or_grid_fire2 <- or_grid_fire2 %>%
    mutate(n_fires_tot = rowSums(across(starts_with("n_fires"))))

head(or_grid_fire2 %>%
    arrange(desc(n_fires_mp))
)

head(or_grid_fire2 %>%
    arrange(desc(n_fires_tot))
)

# max number of mp fires: 9
# max number of total fires: 109
```

```{r}
# n fires
or_grid_fire2 %>%
    ggplot(aes(geometry=geometry, fill=n_fires_tot)) +
    geom_sf() +
    scale_fill_viridis(option="inferno") +
    theme_void()

# n fires is log transformed!
log_n_fires <- or_grid_fire2 %>%
    ggplot(aes(geometry=geometry, fill=n_fires_tot)) +
    geom_sf() +
    scale_fill_viridis(option="inferno", trans="log", na.value="black") +
    theme_void()

log_n_fires

log_n_fires_ind <- or_grid_fire2 %>%
    ggplot(aes(geometry=geometry, fill=n_fires_tot, alpha=is_indigenous)) +
    geom_sf() +
    scale_alpha_binned(range=c(0.6, 1)) + 
    scale_fill_viridis(option="inferno", trans="log", na.value="black") +
    theme_void()

log_n_fires_ind

ggsave("maps/log_n_fires_indigenous.jpeg", log_n_fires_ind)
```

### Adding fire identifier (MP)

```{r}
# getting grid index as list
grid_index <- or_grid_df$index

# new tibble that doesn't have a matrix
fire_tibble2 = cbind(tibble(data.frame(fire_mp_int)), grid_index)
names(fire_tibble2)

# for every grid index and fire, shows the unique fire identifier (unique_id) and whether that fire was in that grid cell (is_fire)
long_fire <- fire_tibble2 %>%
    pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") %>%
    mutate(fire_index = as.numeric(str_remove(fire_index, "X"))) %>%
    mutate(
        unique_id = fire_mp$UNQE_FI[fire_index],
        acres = fire_mp$GIS_ACR[fire_index],
        fire_management_n = fire_mp$fr_mgm_[fire_index],
        duration_hrs = fire_mp$drtn_hr[fire_index]
    )

# confirming the order didn't shift around for some reason
sum(((long_fire %>% pull(unique_id))[1:100] == (fire_mp %>% pull(UNQE_FI))[1:100])==FALSE)

# creating new limited id column that has unique identifier if fire is present in that square, otherwise NA
# same for acres, fire_management_n, duration_hrs
# then pivoting so the unique ID shows in grid index col where the fire is present
wide_fire <- long_fire %>%
    # mutating unique ID to "missing" if NA and adding the constructed fire_index for referencing
    mutate(unique_id = ifelse(is.na(unique_id), paste("MISSING", as.symbol(fire_index)), unique_id)) %>%
    mutate(
        limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
    ) %>%
    pivot_wider(names_from = grid_index, values_from = limited_uid)

head(wide_fire)

# creating new tibble with grid index
# empty columns for unique identifiers, acres, mgmt complexity, duration 
fire_lst_tibblemp <- tibble(grid_index)
head(fire_lst_tibblemp,5)

start_index = which(1==names(wide_fire))-1

# for every grid cell, pulling unique fire identifiers for fires that occurred in that cell, adding them to fire_lst_tibble
for (x in 1:max(grid_index)) {
    # pull list of NAs and fire identifiers for each col/grid cell
    full_lst <- pull(wide_fire, x+start_index)
    # list of indexes for present fires
    fire_idx <- which(!is.na(full_lst))
    # list of present fire identifiers
    lst <- full_lst[!is.na(full_lst)]
    # if no fires, NA or 0 as applicable
    if (length(lst) == 0) {
        fire_lst_tibblemp$fire_lstmp[x] = NA
        fire_lst_tibblemp$max_acresmp[x] = 0
        fire_lst_tibblemp$min_mgmtmp[x] = NA
        fire_lst_tibblemp$max_durationmp[x] = 0
    # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
    } else {
        fire_lst_tibblemp$fire_lstmp[x] = paste(unlist(lst), collapse=", ")
        fire_lst_tibblemp$max_acresmp[x] = max(wide_fire$acres[fire_idx], na.rm=TRUE)
        fire_lst_tibblemp$min_mgmtmp[x] = min(wide_fire$fire_management_n[fire_idx], na.rm=TRUE)
        fire_lst_tibblemp$max_durationmp[x] = max(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
    }
}

str(fire_lst_tibblemp)

fire_lst_tibblemp <- fire_lst_tibblemp %>%
    mutate(
        min_mgmtmp = na_if(min_mgmtmp, Inf),
        max_durationmp = na_if(max_durationmp, -Inf)
    )

fire_lst_tibblemp %>% filter(!is.na(fire_lstmp))

# checking everything accurate
ex47 <- pull(wide_fire, `47`)
ex47_idx <- which(!is.na(ex47))
max(wide_fire$duration_hrs[ex47_idx])
ex47 <- ex47[!is.na(ex47)]
ex47
fire_lst_tibblemp[47:50,]

ex47 == fire_lst_tibblemp$fire_lstmp[47]
```

```{r}
# joining or_grid_df and fire_lst_tibble on the grid index
or_grid_firemp <- tibble(left_join(or_grid_fire2, fire_lst_tibblemp, by=c("index"="grid_index")))

or_grid_firemp
names(or_grid_firemp)
```

### Adding fire identifier (PT)

```{r}
# new tibble that doesn't have a matrix
fire_tibble2pt = cbind(tibble(data.frame(fire_pt_int)), grid_index)
names(fire_tibble2pt)

# for every grid index and fire, shows the unique fire identifier (unique_id) and whether that fire was in that grid cell (is_fire)
long_firept <- fire_tibble2pt %>%
    pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") %>%
    mutate(fire_index = as.numeric(str_remove(fire_index, "X"))) %>%
    mutate(
        unique_id = fire_pt$UNQE_FI[fire_index],
        acres = fire_pt$Clclt_A[fire_index],
        fire_management_n = fire_pt$fr_mgm_[fire_index],
        duration_hrs = fire_pt$drtn_hr[fire_index],
        acres_tribal = fire_pt$AcrsTrb[fire_index]
    )

# confirming the order didn't shift around for some reason
(long_firept %>% pull(unique_id))[1:100] == (fire_pt %>% pull(UNQE_FI))[1:100]

# creating new limited id column that has unique identifier if fire is present in that square, otherwise NA
# same for acres, fire_management_n, duration_hrs
# then pivoting so the unique ID shows in grid index col where the fire is present
wide_firept <- long_firept %>%
    # mutating unique ID to "missing" if NA and adding the constructed fire_index for referencing
    mutate(unique_id = ifelse(is.na(unique_id), paste("MISSING", as.symbol(fire_index)), unique_id)) %>%
    mutate(
        limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
    ) %>%
    pivot_wider(names_from = grid_index, values_from = limited_uid)

head(wide_firept)

# creating new tibble with grid index
# empty columns for unique identifiers, acres, mgmt complexity, duration 
fire_lst_tibblept <- tibble(grid_index)
head(fire_lst_tibblept,5)

start_indexpt = which(1==names(wide_firept))-1

# for every grid cell, pulling unique fire identifiers for fires that occurred in that cell, adding them to fire_lst_tibble
for (x in 1:max(grid_index)) {
    # pull list of NAs and fire identifiers for each col/grid cell
    full_lst <- pull(wide_firept, x+start_indexpt)
    # list of indexes for present fires
    fire_idx <- which(!is.na(full_lst))
    # list of present fire identifiers
    lst <- full_lst[!is.na(full_lst)]
    # if no fires, NA or 0 as applicable
    if (length(lst) == 0) {
        fire_lst_tibblept$fire_lstpt[x] = NA
        fire_lst_tibblept$max_acrespt[x] = 0
        fire_lst_tibblept$min_mgmtpt[x] = NA
        fire_lst_tibblept$max_durationpt[x] = 0
        fire_lst_tibblept$max_acres_tribal[x] = 0
    # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
    } else {
        fire_lst_tibblept$fire_lstpt[x] = paste(unlist(lst), collapse=", ")
        fire_lst_tibblept$max_acrespt[x] = max(wide_firept$acres[fire_idx], na.rm=TRUE)
        fire_lst_tibblept$min_mgmtpt[x] = min(wide_firept$fire_management_n[fire_idx], na.rm=TRUE)
        fire_lst_tibblept$max_durationpt[x] = max(wide_firept$duration_hrs[fire_idx], na.rm=TRUE)
        fire_lst_tibblept$max_acres_tribal[x] = max(wide_firept$acres_tribal[fire_idx], na.rm=TRUE)
    }
}

str(fire_lst_tibblept)

fire_lst_tibblept <- fire_lst_tibblept %>%
    mutate(
        min_mgmtpt = na_if(min_mgmtpt, Inf),
        max_durationpt = na_if(max_durationpt, -Inf)
    )

fire_lst_tibblept %>% filter(!is.na(fire_lstpt))

# checking everything accurate
ex20 <- pull(wide_firept, `20`)
ex20_idx <- which(!is.na(ex20))
max(wide_firept$duration_hrs[ex20_idx])
ex20 <- ex20[!is.na(ex20)]
ex20
fire_lst_tibblept[15:20,]

ex20 == fire_lst_tibblept$fire_lstpt[20]
```

```{r}
# joining or_grid_df and fire_lst_tibble on the grid index
or_grid_firemp_pt <- tibble(left_join(or_grid_firemp, fire_lst_tibblept, by=c("index"="grid_index")))

or_grid_firemp_pt
names(or_grid_firemp_pt)
```

### Combined MP/PT stats

```{r}
or_grid_firemp_pt <- or_grid_firemp_pt %>%
    unite(fire_lst, fire_lstmp, fire_lstpt, na.rm=TRUE, remove=FALSE, sep=", ") %>%
    mutate(fire_lst = ifelse(fire_lst=="", NA, fire_lst)) %>%
    rowwise() %>%
    mutate(
        max_acres = max(max_acresmp, max_acrespt, na.rm=TRUE),
        min_mgmt = min(min_mgmtmp, min_mgmtmp, na.rm=TRUE),
        max_duration = max(max_durationmp, max_durationpt, na.rm=TRUE)
    ) 

or_grid_firemp_pt <- or_grid_firemp_pt %>%
    mutate(
        min_mgmt = na_if(min_mgmt, min_mgmt==Inf),
        max_duration = na_if(max_duration, max_duration==-Inf)
    ) %>%
    mutate(max_size_class = cut(
        max_acres, 
        breaks=c(-0.01, 0.25, 9.99, 99.99, 299.99, 999.99, 4999.99, 9000000),
        labels = c("Class A", "Class B", "Class C", "Class D", "Class E", "Class F", "Class G")
    )) 

str(or_grid_firemp_pt)

```

```{r}
# test visualization
or_grid_firemp_pt %>%
    ggplot(aes(geometry=geometry, fill=max_acres)) +
    geom_sf(linewidth=0.1) +
    scale_fill_viridis(option="inferno") +
    theme_void()

size_class_plt <- or_grid_firemp_pt %>%
    mutate(max_size_class = fct_reorder(max_size_class, max_acres)) %>%
    ggplot(aes(geometry=geometry, fill=max_size_class)) +
    geom_sf(linewidth=0.1) +
    scale_fill_viridis_d(option="inferno") +
    theme_void()

size_class_plt

ggsave("maps/fire_size_class_new.jpeg", size_class_plt)
```

```{r}
or_grid_firemp_pt %>%
    group_by(is_indigenous) %>%
    summarize(n=n())
```

## Adding InFORM/IFPH prescribed burn data

### Adding fire count

```{r}
# fire tibble with whether each fire occurred in each grid cell
rx_mp_int <- st_intersects(or_grid, rx_fire_mp, sparse=FALSE)
rx_mp_tibble <- tibble(rx_mp_int)
head(rx_mp_tibble)
dim(rx_mp_tibble)
dim(rx_mp_tibble$rx_mp_int)

# fire tibble for the point geometries
rx_pt_int <- st_contains(or_grid, rx_fire_pt, sparse=FALSE)
rx_pt_tibble <- tibble(rx_pt_int)
dim(rx_pt_tibble)
dim(rx_pt_tibble$rx_pt_int)

# combining fire_tibble and or_grid_df to get counts of fires
or_grid_rx_mp <- cbind(or_grid_firemp_pt, rx_mp_tibble)

# creating sum of fires across each row/grid cell
or_grid_rx_mp <- or_grid_rx_mp %>%
    mutate(across(starts_with("rx"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_rx_mp = rowSums(across(starts_with("rx")))
    )

# removing the interim fire columns
or_grid_rx_mp <- or_grid_rx_mp %>% select(-rx_mp_int)

or_grid_rx_pt <- cbind(or_grid_rx_mp, rx_pt_tibble)

or_grid_rx_pt <- or_grid_rx_pt %>%
    mutate(across(starts_with("rx"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_rx_pt = rowSums(across(starts_with("rx")))
    )

or_grid_rx_pt <- or_grid_rx_pt %>% select(-rx_pt_int)

names(or_grid_rx_pt)

or_grid_rx_pt <- or_grid_rx_pt %>%
    mutate(n_rx_tot = rowSums(across(starts_with("n_rx"))))

head(or_grid_rx_pt %>%
    arrange(desc(n_rx_tot))
)

# max number of rx fires: 8
```

```{r}
rx_fires_plt <- or_grid_rx_pt %>%
    ggplot(aes(geometry=geometry, fill=n_rx_tot)) +
    geom_sf(color="black") +
    scale_fill_viridis(option="inferno") +
    theme_minimal()

rx_fires_plt

ggsave("maps/rx_fires_new.jpeg", rx_fires_plt)

rx_fires_ind <- or_grid_rx_pt %>%
    ggplot(aes(geometry=geometry, fill=n_rx_tot, alpha=is_indigenous)) +
    geom_sf() +
    scale_alpha_binned(range=c(0.6, 1)) + 
    scale_fill_viridis(option="inferno") +
    theme_minimal()

rx_fires_ind

ggsave("maps/rx_fires_indigenous_new.jpeg", rx_fires_ind)
```

### Adding RX fire identifier (MP)

```{r}
# there is only one RX fire in the multipolygon data, so we can add it manually
# in a world where there were more, we would repeat the process from the fires
or_grid_rx_pt %>%
    filter(n_rx_mp > 0)

rx_fire_mp

# creating new tibble with grid index
rx_grid_idmp <- (or_grid_rx_pt %>% filter(n_rx_mp > 0))$index
rx_lstmp <- rx_fire_mp$UNQE_FI
rx_max_acresmp <- rx_fire_mp$GIS_ACR
rx_min_mgmtmp <- rx_fire_mp$fr_mgm_
rx_max_durationmp <- rx_fire_mp$drtn_hr
rx_lst_tibblemp <- tibble(rx_grid_idmp, rx_lstmp, rx_max_acresmp, rx_min_mgmtmp, rx_max_durationmp)

rx_lst_tibblemp
```

```{r}
# joining or_grid_df and fire_lst_tibble on the grid index
or_grid_rxmp <- tibble(left_join(or_grid_rx_pt, rx_lst_tibblemp, by=c("index"="rx_grid_idmp")))

or_grid_rxmp
names(or_grid_rxmp)
```

### Adding RX fire identifier (PT)

```{r}
# new tibble that doesn't have a matrix
rx_tibble2pt = cbind(tibble(data.frame(rx_pt_int)), grid_index)
names(rx_tibble2pt)

# for every grid index and fire, shows the unique fire identifier (unique_id) and whether that fire was in that grid cell (is_fire)
long_rxpt <- rx_tibble2pt %>%
    pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") %>%
    mutate(fire_index = as.numeric(str_remove(fire_index, "X"))) %>%
    mutate(
        unique_id = rx_fire_pt$UNQE_FI[fire_index],
        acres = rx_fire_pt$Clclt_A[fire_index],
        fire_management_n = rx_fire_pt$fr_mgm_[fire_index],
        duration_hrs = rx_fire_pt$drtn_hr[fire_index],
        acres_tribal = rx_fire_pt$AcrsTrb[fire_index]
    )

# confirming the order didn't shift around for some reason
(long_rxpt %>% pull(unique_id))[1:100] == (rx_fire_pt %>% pull(UNQE_FI))[1:100]

# creating new limited id column that has unique identifier if fire is present in that square, otherwise NA
# same for acres, fire_management_n, duration_hrs
# then pivoting so the unique ID shows in grid index col where the fire is present
wide_rxpt <- long_rxpt %>%
    # mutating unique ID to "missing" if NA and adding the constructed fire_index for referencing
    mutate(unique_id = ifelse(is.na(unique_id), paste("MISSING", as.symbol(fire_index)), unique_id)) %>%
    mutate(
        limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
    ) %>%
    pivot_wider(names_from = grid_index, values_from = limited_uid)

head(wide_rxpt)

# creating new tibble with grid index
# empty columns for unique identifiers, acres, mgmt complexity, duration 
rx_lst_tibblept <- tibble(grid_index)
head(rx_lst_tibblept,5)

start_indexrxpt = which(1==names(wide_rxpt))-1

# for every grid cell, pulling unique fire identifiers for fires that occurred in that cell, adding them to fire_lst_tibble
for (x in 1:max(grid_index)) {
    # pull list of NAs and fire identifiers for each col/grid cell
    full_lst <- pull(wide_rxpt, x+start_indexrxpt)
    # list of indexes for present fires
    fire_idx <- which(!is.na(full_lst))
    # list of present fire identifiers
    lst <- full_lst[!is.na(full_lst)]
    # if no fires, NA or 0 as applicable
    if (length(lst) == 0) {
        rx_lst_tibblept$rx_lstpt[x] = NA
        rx_lst_tibblept$rx_max_acrespt[x] = 0
        rx_lst_tibblept$rx_min_mgmtpt[x] = NA
        rx_lst_tibblept$rx_max_durationpt[x] = 0
        rx_lst_tibblept$rx_max_acres_tribal[x] = 0
    # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
    } else {
        rx_lst_tibblept$rx_lstpt[x] = paste(unlist(lst), collapse=", ")
        rx_lst_tibblept$rx_max_acrespt[x] = max(wide_rxpt$acres[fire_idx], na.rm=TRUE)
        rx_lst_tibblept$rx_min_mgmtpt[x] = min(wide_rxpt$fire_management_n[fire_idx], na.rm=TRUE)
        rx_lst_tibblept$rx_max_durationpt[x] = max(wide_rxpt$duration_hrs[fire_idx], na.rm=TRUE)
        rx_lst_tibblept$rx_max_acres_tribal[x] = max(wide_rxpt$acres_tribal[fire_idx], na.rm=TRUE)
    }
}

rx_lst_tibblept <- rx_lst_tibblept %>%
    mutate(
        rx_min_mgmtpt = na_if(rx_min_mgmtpt, Inf),
        rx_max_durationpt = na_if(rx_max_durationpt, -Inf)
    )

rx_lst_tibblept %>%
    filter(!is.na(rx_lstpt))

# checking everything accurate
ex308 <- pull(wide_rxpt, `308`)
ex308_idx <- which(!is.na(ex308))
max(wide_rxpt$duration_hrs[ex308_idx])
ex308 <- ex308[!is.na(ex308)]
ex308
rx_lst_tibblept[308,]

ex308 == rx_lst_tibblept$rx_lstpt[308]
```

```{r}
# joining or_grid_df and fire_lst_tibble on the grid index
or_grid_all_fire <- tibble(left_join(or_grid_rxmp, rx_lst_tibblept, by=c("index"="grid_index")))

or_grid_all_fire
names(or_grid_all_fire)
```

### Combined MP/PT stats for RX

```{r}
or_grid_all_fire <- or_grid_all_fire %>%
    unite(rx_lst, rx_lstmp, rx_lstpt, na.rm=TRUE, remove=FALSE, sep=", ") %>%
    mutate(rx_lst = na_if(rx_lst, "")) %>%
    rowwise() %>%
    mutate(
        rx_max_acres = max(rx_max_acresmp, rx_max_acrespt, na.rm=TRUE),
        rx_min_mgmt = min(rx_min_mgmtmp, rx_min_mgmtmp, na.rm=TRUE),
        rx_max_duration = max(rx_max_durationmp, rx_max_durationpt, na.rm=TRUE)
    ) 

or_grid_all_fire <- or_grid_all_fire %>%
    mutate(
        rx_min_mgmt = na_if(rx_min_mgmt, Inf),
        rx_max_duration = na_if(rx_max_duration, -Inf)
    ) %>%
    mutate(rx_max_size_class = cut(
        rx_max_acres, 
        breaks=c(-0.01, 0.25, 9.99, 99.99, 299.99, 999.99, 4999.99, 9000000),
        labels = c("Class A", "Class B", "Class C", "Class D", "Class E", "Class F", "Class G")
    ))

str(or_grid_all_fire)

```

# Adding GBIF data

## Read in and make sf

```{r}
gbif <- read_csv("data/new_gbif_dataset.csv")
```

```{r}
gbif_sf <- st_as_sf(gbif, coords=c("decimalLongitude", "decimalLatitude"), crs="WGS84")
gbif_sf <- gbif_sf %>% select(-`...1`)
```

```{r}
str(gbif_sf)

gbif_sf <- gbif_sf %>%
    mutate(
        species = as_factor(species),
        eventDate = ifelse(str_starts(eventDate, "20"), ymd_hms(eventDate, truncated=3), mdy_hms(eventDate, truncated=3)),
        basisOfRecord = as_factor(basisOfRecord),
        license = as_factor(license)
    ) 

```

# Species counts

```{r}
unique(gbif_sf$species)

# datasets filtered for each species
qg_gbif <- gbif_sf %>%
    filter(species == "Quercus garryana")

xt_gbif <- gbif_sf %>%
    filter(species == "Xerophyllum tenax")

tp_gbif <- gbif_sf %>%
    filter(species == "Thuja plicata")

pt_gbif <- gbif_sf %>%
    filter(species == "Populus trichocarpa")

ru_gbif <- gbif_sf %>%
    filter(species == "Rubus ursinus")

pp_gbif <- gbif_sf %>%
    filter(species == "Pinus ponderosa")

ss_gbif <- gbif_sf %>%
    filter(species == "Salix scouleriana")

```

```{r}
# contains lists/tibbles for each species
qg_con <- st_contains(or_grid, qg_gbif, sparse=FALSE)
qg_tibble <- tibble(qg_con)

xt_con <- st_contains(or_grid, xt_gbif, sparse=FALSE)
xt_tibble <- tibble(xt_con)

tp_con <- st_contains(or_grid, tp_gbif, sparse=FALSE)
tp_tibble <- tibble(tp_con)

pt_con <- st_contains(or_grid, pt_gbif, sparse=FALSE)
pt_tibble <- tibble(pt_con)

ru_con <- st_contains(or_grid, ru_gbif, sparse=FALSE)
ru_tibble <- tibble(ru_con)

pp_con <- st_contains(or_grid, pp_gbif, sparse=FALSE)
pp_tibble <- tibble(pp_con)

ss_con <- st_contains(or_grid, ss_gbif, sparse=FALSE)
ss_tibble <- tibble(ss_con)

head(qg_tibble)
```

```{r}
# adding counts per cell to overall df
# quercus garryana
or_grid_gbif <- cbind(or_grid_all_fire, qg_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("qg_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_quercus = rowSums(across(starts_with("qg_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-qg_con)

# xerophyllum tenax
or_grid_gbif <- cbind(or_grid_gbif, xt_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("xt_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_xerophyllum = rowSums(across(starts_with("xt_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-xt_con)

# thuja plicata
or_grid_gbif <- cbind(or_grid_gbif, tp_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("tp_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_thuja = rowSums(across(starts_with("tp_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-tp_con)

# populus trichocarpa
or_grid_gbif <- cbind(or_grid_gbif, pt_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("pt_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_populus = rowSums(across(starts_with("pt_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-pt_con)

# rubus ursinus
or_grid_gbif <- cbind(or_grid_gbif, ru_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("ru_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_rubus = rowSums(across(starts_with("ru_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-ru_con)

# pinus ponderosa
or_grid_gbif <- cbind(or_grid_gbif, pp_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("pp_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_pinus = rowSums(across(starts_with("pp_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-pp_con)

# salix scouleriana
or_grid_gbif <- cbind(or_grid_gbif, ss_tibble)

or_grid_gbif <- or_grid_gbif %>%
    mutate(across(starts_with("ss_con"), ~ ifelse(.x==TRUE, 1, 0))) %>% 
    mutate(
        n_salix = rowSums(across(starts_with("ss_con")))
    )

or_grid_gbif <- or_grid_gbif %>% select(-ss_con)

names(or_grid_gbif)
```

```{r}
# adding plant sums

startid <- which("n_quercus"==names(or_grid_gbif))
startid
endid <- which("n_salix"==names(or_grid_gbif))
endid

or_grid_gbif <- or_grid_gbif %>% 
    mutate(plant_sum = rowSums(or_grid_gbif[, startid:endid]))

```

# Historic Vegetation info

```{r}
hist_veg <- st_read("../Historic_Vegetation_84513795379117550/Historic_Vegetation.shp")
hist_veg <- st_transform(hist_veg, crs="WGS84")
```

## adding historic vegetation cols of interest

```{r}
# datasets filtered to each species
hist_pinus <- hist_veg %>%
    filter(VEGNAME %in% c("Ponderosa pine", "Ponderosa pine-lodgepole pine","Douglas fir-ponderosa pine"))

hist_salix <- hist_veg %>%
    filter(VEGNAME=="Willows")

hist_quercus <- hist_veg %>%
    filter(VEGNAME %in% c("White oak woodland", "White oak-black oak"))

hist_populus <- hist_veg %>%
    filter(VEGNAME=="Black cottonwood riparian woodland")

hist_thuja <- hist_veg %>%
    filter(VEGNAME=="Western redcedar")

```

```{r}
# contains lists/tibbles for each species
hist_pinus_grid <- st_intersects(or_grid, hist_pinus, sparse=FALSE)
hist_pinus_tibble <- tibble(hist_pinus_grid)

hist_salix_grid <- st_intersects(or_grid, hist_salix, sparse=FALSE)
hist_salix_tibble <- tibble(hist_salix_grid)

hist_quercus_grid <- st_intersects(or_grid, hist_quercus, sparse=FALSE)
hist_quercus_tibble <- tibble(hist_quercus_grid)

hist_populus_grid <- st_intersects(or_grid, hist_populus, sparse=FALSE)
hist_populus_tibble <- tibble(hist_populus_grid)

hist_thuja_grid <- st_intersects(or_grid, hist_thuja, sparse=FALSE)
hist_thuja_tibble <- tibble(hist_thuja_grid)

```

```{r}
# adding presence by cell to overall df
# pinus
or_grid_histveg <- cbind(or_grid_gbif, hist_pinus_tibble)

or_grid_histveg <- or_grid_histveg %>%
    mutate(across(starts_with("hist_pinus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_pinus = ifelse(rowSums(across(starts_with("hist_pinus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_pinus_grid)

# salix
or_grid_histveg <- cbind(or_grid_histveg, hist_salix_tibble)

or_grid_histveg <- or_grid_histveg %>%
    mutate(across(starts_with("hist_salix_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_salix = ifelse(rowSums(across(starts_with("hist_salix_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_salix_grid)

# quercus
or_grid_histveg <- cbind(or_grid_histveg, hist_quercus_tibble)

or_grid_histveg <- or_grid_histveg %>%
    mutate(across(starts_with("hist_quercus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_quercus = ifelse(rowSums(across(starts_with("hist_quercus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_quercus_grid)

# populus 
or_grid_histveg <- cbind(or_grid_histveg, hist_populus_tibble)

or_grid_histveg <- or_grid_histveg %>%
    mutate(across(starts_with("hist_populus_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_populus = ifelse(rowSums(across(starts_with("hist_populus_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_populus_grid)

# thuja
or_grid_histveg <- cbind(or_grid_histveg, hist_thuja_tibble)

or_grid_histveg <- or_grid_histveg %>%
    mutate(across(starts_with("hist_thuja_grid"), ~ ifelse(.x==TRUE, 1, 0))) %>%
    mutate(
        hist_thuja = ifelse(rowSums(across(starts_with("hist_thuja_grid"))) == 0, 0, 1)
    ) %>%
    select(-hist_thuja_grid) 

names(or_grid_histveg)
```

## summing everything (GBIF and historic)

```{r}
or_grid_histveg <- or_grid_histveg %>%
    mutate(
        sum_pinus = rowSums(across(ends_with("pinus"))),
        sum_salix = rowSums(across(ends_with("salix"))),
        sum_quercus = rowSums(across(ends_with("quercus"))),
        sum_populus = rowSums(across(ends_with("populus"))),
        sum_thuja = rowSums(across(ends_with("thuja"))),
    ) %>% 
    mutate(
        total_plant_sum = rowSums(across(c(starts_with("sum"), n_xerophyllum, n_rubus)))
    )

names(or_grid_histveg)
```

# finalizing join

```{r}
class(or_grid_histveg)
verbose_grid_shp <- st_as_sf(or_grid_histveg, crs="WGS84")
object.size(verbose_grid_shp)
st_write(verbose_grid_shp, "data/processed_grid/verbose_grid.shp", append=FALSE)

object.size(or_grid_histveg)
write_csv(or_grid_histveg, "data/processed_grid/verbose_grid.csv")
```

```{r}
names(or_grid_histveg)
processed_grid <- or_grid_histveg %>%
    select(
        geometry,
        index, 
        is_indigenous,
        n_fires_tot,
        fire_lst,
        max_acres, 
        max_acres_tribal,
        min_mgmt, 
        max_duration,
        max_size_class,
        n_rx_tot,
        rx_lst,
        rx_max_acres,
        rx_max_acres_tribal,
        rx_min_mgmt,
        rx_max_duration,
        rx_max_size_class,
        sum_pinus,
        sum_salix,
        sum_quercus,
        sum_populus,
        sum_thuja, 
        n_xerophyllum,
        n_rubus,
        total_plant_sum
    ) %>%
    rename(
        n_wildfires = n_fires_tot,
        wf_uids = fire_lst,
        n_rx_fires = n_rx_tot,
        rx_uids = rx_lst,
        n_pinus = sum_pinus,
        n_salix = sum_salix,
        n_quercus = sum_quercus,
        n_populus = sum_populus,
        n_thuja = sum_thuja, 
        plant_sum = total_plant_sum
    )

names(processed_grid)

# i kind of thought I had removed all of these already. but evidently I had not!
processed_grid <- processed_grid %>%
    mutate(
        min_mgmt = na_if(min_mgmt, Inf),
        max_duration = na_if(max_duration, -Inf)
    )

```

```{r}
class(processed_grid)
processed_grid_shp <- st_as_sf(processed_grid, crs="WGS84")
object.size(processed_grid_shp)
st_write(processed_grid_shp, "data/processed_grid/processed_grid.shp", append=FALSE)

object.size(processed_grid)
write_csv(processed_grid, "data/processed_grid/processed_grid.csv")

```