---
title: "IFPH InFORM Join"
author: "Spencer Fiden"
date: "2026-05-29"
output: html_document
---

Joining IFPH and InFORM information into two files (one with point geometry and one with multipolygon geometry). Reads from local system for InFORM, from `data/fire_df_shp` for IFPH. 

Writes several files! Writes multipolygon (IFPH data inner joined with InFORM to include both records) to `data/ifph_inform_multipolygon`. Writes point (InFORM data with present lat/lon in Oregon as coordinate points) to `data/ifph_inform_point`. Writes a complete csv with all entries to `data/ifph_inform`. 

# IFPH InFORM join

## Ingestion from download

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
```

```{r}
# reading in cleaned files
ifph <- st_read("data/fire_df_shp/fire_shp.shp")

# inform stored locally due to github size constraints
inform <- read_csv("../DATA-510/fire_df.csv")

```

```{r}
dim(inform) # 304579 rows
dim(ifph) # 1683 rows
```

# Joining them

## Checking Unique Fire ID for join validity

```{r}
# making tibble for this so it runs faster
tibble(ifph) %>%
    mutate(FORID = str_remove_all(FORID, "[{}]")) %>%
    mutate(
        unqe_match = ifelse(UNQE_FIRE_ %in% inform$UniqueFireIdentifier, T, F),
        forid_match = ifelse(FORID %in% inform$FORID, T, F),
        globalid_match = ifelse(GlobalID %in% inform$GlobalID, T, F)
    ) %>%
    group_by(unqe_match, forid_match, globalid_match)%>%
    summarize(n=n())

# FORID and GlobalID have no matches, will be using just unique fire identifier

# seem to have compatible formats anyway - unclear why they don't match
head(inform %>% select(UniqueFireIdentifier, FORID, GlobalID))
head(ifph %>% select(UNQE_FIRE_, FORID, GlobalID))

sum(tibble(ifph)$UNQE_FIRE_ %in% inform$UniqueFireIdentifier)
# 817 IFPH records have matches in InFORM
```

```{r}
tibble(ifph) %>%
    group_by(UNQE_FIRE_) %>%
    summarize(n=n()) %>%
    filter(n > 1) %>%
    arrange(desc(n))

inform %>%
    group_by(UniqueFireIdentifier) %>%
    summarize(n=n()) %>%
    filter(n > 1, !is.na(UniqueFireIdentifier)) %>%
    arrange(desc(n))

# everything with duplicated unique fire identifiers has a distinct global id for each entry
```

```{r}
ifph_dup <- tibble(ifph) %>%
    group_by(UNQE_FIRE_) %>%
    summarize(n=n()) %>%
    filter(n > 1) %>%
    arrange(desc(n))

ifph %>%
    filter(UNQE_FIRE_ %in% ifph_dup$UNQE_FIRE_[2]) %>%
    #select(-c(IRWINID, FORID, INCIDENT)) %>%
    #arrange(UNQE_FIRE_) %>%
    ggplot(aes(geometry=geometry)) +
    geom_sf() +
    coord_sf() +
    theme_minimal()

ifph %>%
    filter(UNQE_FIRE_ %in% ifph_dup$UNQE_FIRE_[3]) %>%
    select(-c(IRWINID, FORID, INCIDENT)) %>%
    arrange(UNQE_FIRE_)

```

The duplicates in IFPH seem to be due to having multiple unconnected polygons mapping them. This would actually explain the differences between IFPH and InFORM acreage numbers! Probably, in which case we did make the right choice opting for IFPH numbers, as it'll only be the acreage in a single cell.

```{r}
inform_dup <- inform %>%
    group_by(UniqueFireIdentifier) %>%
    summarize(n=n()) %>%
    filter(n > 1) %>%
    arrange(desc(n))

inform %>%
    filter(UniqueFireIdentifier %in% inform_dup$UniqueFireIdentifier[1]) %>%
    print(width=Inf)
```

At least some of the duplicate fires in InFORM have different incident names and acres sizes assigned to the same fire identifier. Not completely sure what's up with that. But if we were to choose one, I don't know what we would choose - may just maintain all of them. 

## checking lat lon presence

```{r}
inform %>%
    group_by(Incident.Type.Category) %>%
    summarize(
        lat_missing = sum(is.na(Initial.Latitude)),
        lon_missing = sum(is.na(Initial.Longitude)),
        n=n()
    ) %>%
    mutate(
        pct_present = 1 - (lat_missing/n)
    )

tibble(ifph) %>% 
    mutate(is_ifph = TRUE) %>%
    full_join(inform %>% mutate(is_inform = TRUE) %>% filter(!is.na(UniqueFireIdentifier)), by=c("UNQE_FIRE_" = "UniqueFireIdentifier")) %>%
    group_by(is_ifph, is_inform, Incident.Type.Category) %>%
    summarize(
        n=n(),
        lat_missing = sum(is.na(Initial.Latitude))
    ) %>%
    mutate(
        pct_present = 1 - (lat_missing/n)
    )

```

# Joining on UNQE_FIRE_

```{r}
inform %>%
    filter(UniqueFireIdentifier %in% ifph$UNQE_FIRE_) %>%
    group_by(is.na(Fire.Mgmt.Complexity)) %>% 
    summarize(n=n())
# fire management complexity missing for 649 matched fires, present for 142

tibble(ifph) %>% 
    mutate(is_ifph = TRUE) %>%
    full_join(inform %>% mutate(is_inform = TRUE) %>% filter(!is.na(UniqueFireIdentifier)), by=c("UNQE_FIRE_" = "UniqueFireIdentifier")) %>%
    group_by(is_ifph, is_inform, Calendar.Year) %>%
    summarize(n=n()) %>%
    arrange(desc(Calendar.Year))
# there are a significant number of many to many joins, but this is a full join.
```

```{r}
# inner join on IFPH
# non-IFPH data will be point geometries from inform, IFPH fires without matched inform data will not have our other fire severity features
ifph_inform <- ifph %>% 
    mutate(is_ifph = TRUE) %>%
    inner_join(inform %>% mutate(is_inform = TRUE) %>% filter(!is.na(UniqueFireIdentifier)), by=c("UNQE_FIRE_" = "UniqueFireIdentifier"))

# assorted checks
tibble(ifph_inform) %>% 
    filter(FIRE_YEAR_ >= 2015, FIRE_YEAR_ <= 2025) %>%
    group_by(is_inform, FIRE_YEAR_) %>%
    summarize(n=n()) %>%
    arrange(desc(n)) %>%
    print(n=20)

ifph_inform %>%
    filter(FIRE_YEAR_ != Calendar.Year)
# the year always matches for included fires!

tibble(ifph_inform) %>% 
    group_by(is_inform) %>%
    summarize(
        n=n(),
        quant_year = as_tibble_row(quantile(FIRE_YEAR_))
    ) %>%
    arrange(desc(n)) 

tibble(ifph_inform) %>%
    filter(is.na(Fire.Mgmt.Complexity)) %>%
    summarize(n=n())

tibble(ifph_inform) %>%
    mutate(acre_diff = abs(Calculated.Acres - GIS_ACRES)) %>%
    summarize(
        mean_diff = mean(acre_diff),
        quant_diff = as_tibble_row(quantile(acre_diff))
    )

ifph_inform %>%
    mutate(acre_diff = abs(Calculated.Acres - GIS_ACRES)) %>%
    mutate(area = st_area(geometry)/4047) %>%
    filter(acre_diff > 10) %>%
    arrange(desc(acre_diff)) %>%
    select(GIS_ACRES, Calculated.Acres, acre_diff, area, FIRE_YEAR_)
# some significant differences here. not ideal. 
# GIS_ACRES align with calculation based on multipolygon
```

# Adding coordinate data from inform records that don't match ifph

```{r}
tibble(ifph_inform) %>%
    group_by(is_inform) %>%
    summarize(n=n())

inform %>%
    mutate(unqe_match = UniqueFireIdentifier %in% ifph_inform$UNQE_FIRE_) %>%
    group_by(unqe_match) %>%
    summarize(n=n()) # so i'm not sure why this is 791 and not 819? seems strange.

# filtering to fire IDs not present in UNQE_FIRE_
# selecting only columns we will use in our analysis
inform_lim <- inform_or %>% 
    filter(!(UniqueFireIdentifier %in% ifph_inform$UNQE_FIRE_)) %>% 
    select(UniqueFireIdentifier, Calculated.Acres, size_class, latest_out, duration_hrs, Fire.Mgmt.Complexity, fire_mgmt_n, Incident.Type.Category, AcresTribal, Calendar.Year, geometry)

# selecting only matching cols
ifph_inform_lim <- ifph_inform %>%
    select(GIS_ACRES, Calculated.Acres, UNQE_FIRE_, FIRE_YEAR_, is_ifph, size_class.x, latest_out, duration_hrs, Fire.Mgmt.Complexity, fire_mgmt_n, Incident.Type.Category, AcresTribal, Calendar.Year, is_inform, geometry)

# renaming columns to match, adding NAs for cols that are ifph-specifc
inform_lim <- inform_lim %>%
    rename(c(
        UNQE_FIRE_ = UniqueFireIdentifier,
    )) %>%
    mutate(
        GIS_ACRES = NA,
        FIRE_YEAR_ = NA,
        is_ifph = FALSE,
        is_inform = TRUE
    )

ifph_inform_lim <- ifph_inform_lim %>%
    rename(size_class = size_class.x)

# ensuring names all match
names(inform_lim) %in% names(ifph_inform_lim)
names(ifph_inform_lim) %in% names(inform_lim)

# checking dims - everything lines up
dim(inform_lim)
dim(ifph_inform_lim)
dim(rbind(inform_lim, ifph_inform_lim))

# creating full version with rbind
ifph_inform_full <- rbind(inform_lim, ifph_inform_lim)

# so it turns out. you can't write multipolygon and point geometries to the same shapefile. so it'll just have to be two shapefiles. whatever.

```

# Writing to shp/csv

```{r}
object.size(ifph_inform_lim)
st_write(ifph_inform_lim, "data/ifph_inform_multipolygon/ifph_inform_multipolygon.shp", append=FALSE)

object.size(inform_lim)
st_write(inform_lim, "data/ifph_inform_point/ifph_inform_point.shp", append=FALSE)

ifph_inform_full_df <- tibble(ifph_inform_full) %>% select(-geometry)

object.size(ifph_inform_full_df)

write_csv(ifph_inform_full_df, "data/ifph_inform/ifph_inform_full.csv")

```
