---
title: "Create time bound grid!"
author: "Spencer Fiden"
date: "2026-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# importing libraries
library(tidyverse)
library(ggplot2)
library(sf)
library(viridis)

```

```{r}
og_grid <- st_read("data/processed_grid/processed_grid.shp")
```

A lot of our information is not time bound - ecology data from GBIF and Historic Vegetation and Indigenous land data from LandMark are not timebound 

(Note: technically GBIF sightings are timestamped but for our purposes we are going to assume that our plants have not significantly changed over these ten years, as we are using that data moreso to get a sense of the regions ecology wrt the presence of the seven species we're investigating). 

Therefore, we're going to remove all of the fire data from the grid and start from there as our baseline. We will have 10 entries per grid cell, one for each year we are including in our project. We will then populate the grid with fires based on the year they occurred in. 

# Base grid

```{r}
names(og_grid)
# removing fire related columns
grid_base <- og_grid %>%
    select(-c(
        n_wldfr, 
        wf_uids, 
        max_crs, 
        mx_crs_, 
        mn_mgmt, 
        mx_drtn, 
        mx_sz_c, 
        n_rx_fr, 
        rx_uids, 
        rx_mx_c, 
        rx_mx_c_, 
        rx_mn_m, 
        rx_mx_d, 
        rx_mx_s_
    ))

names(grid_base)
```

Base grid includes index, whether the cell includes indigenous land, counts for our 7 species, plant sum, and geometry columns. 

The new grid will add:

- year
- number of fires that occurred in that year
- number of prescribed burns that occurred in that year 
- severity metrics for wildfires (max duration, max acres burned, max tribal acres burned, min management complexity, size class)
- unique identifiers for fires from that year per cell

```{r}
# setting start and end year
START_YEAR <- 2015
END_YEAR <- 2025
```

```{r}
# creating base for df rows to add
grid_year <- grid_base %>%
    mutate(year = START_YEAR)

# for every year in our time frame, duplicate grid_base and add year column for that year, attach to grid_year df
for (year in (START_YEAR+1):END_YEAR) {
    df <- grid_base %>% mutate(year = year)
    grid_year <- rbind(grid_year, df)
}

tibble(grid_year) %>%
    group_by(year) %>%
    summarize(n=n())

```

```{r}
# creating unique row identifier - year and grid index (with 0s added as needed so that they're all the same length)
max_id_char <- nchar(as.character(max(grid_year$index)))

grid_year <- grid_year %>%
    mutate(row_id = paste0(year, strrep("0", max_id_char-nchar(as.character(index))), index)) 

tail(grid_year$row_id)
```

# Fire data

```{r}
# reading in combined IFPH and InFORM data created in ifph-inform-join
full_fire_mp <- st_read("data/ifph_inform_multipolygon/ifph_inform_multipolygon.shp")

# reading in point data from InFORM lat/lon data (also created in ifph-inform-join)
full_fire_pt <- st_read("data/ifph_inform_point/ifph_inform_point.shp")

# separating into RX and WF
# note that this removes fires that were present in IFPH but not InFORM (we would be missing duration/mgmt complexity info for these as well)
rx_fire_mp <- full_fire_mp %>%
    filter(Inc_T_C == "RX")

fire_mp <- full_fire_mp %>%
    filter(Inc_T_C != "RX")

rx_fire_pt <- full_fire_pt %>%
    filter(Inc_T_C == "RX")

fire_pt <- full_fire_pt %>%
    filter(Inc_T_C != "RX")

```

## Wildfires

### Multipolygon wildfires

```{r}
# grid index - single instance of grid index (not one per year)
grid_index = seq(1:max(grid_base$index))

# initialize empty tibble to hold all values split by year
full_tib_mp <- tibble(
    row_id = numeric(),
    year = numeric(),
    index = numeric()
)

# note: this takes a while to run
# this is not a surprise as it's testing many fires against 10k cells 11 times.
# there may be a more efficient way to do this

for (i in START_YEAR:END_YEAR) {
    print(paste("Year:", i))
    # initialize empty tibble for current year
    fire_lst_tib_mp <- tibble(grid_year) %>% select(row_id, year, index) %>% filter(year==i)

    # intersecting fires by grid for current year
    fire_mp_int <- st_intersects(grid_base, fire_mp %>% filter(Clndr_Y==i), sparse=FALSE)
    # convert to tibble
    fire_mp_tib <- cbind(tibble(data.frame(fire_mp_int)), grid_index)

    # fire_mp with only fires from current year
    fire_mp_curr <- fire_mp %>% filter(Clndr_Y == i)

    # only do following steps if there were fires this year
    if (nrow(fire_mp_curr) > 0) {
        print(paste("N fires:", dim(fire_mp_curr)[1]))

        # long fire with information from every fire in current year
        long_fire <- fire_mp_tib %>%
            pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") 

        # if only 1 fire in a year, index is wrong (can't handle str_remove as labels are different)
        # most will go through
        if (nrow(fire_mp_curr) > 1) {
            long_fire <- long_fire %>%
                mutate(fire_index = as.numeric(str_remove(fire_index, "X"))) 
        } else {
            long_fire <- long_fire %>%
                mutate(fire_index=1)
        }
            
        long_fire <- long_fire %>%
            mutate(
                unique_id = fire_mp_curr$UNQE_FI[fire_index],
                acres = fire_mp_curr$GIS_ACR[fire_index],
                fire_management_n = fire_mp_curr$fr_mgm_[fire_index],
                duration_hrs = fire_mp_curr$drtn_hr[fire_index]
            )
        print(paste("Dim long fire:", dim(long_fire)))

        # wide fire that only includes uid if fire is present in cell
        wide_fire <- long_fire %>%
            mutate(
                limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
            ) %>%
            pivot_wider(names_from = grid_index, values_from = limited_uid)
        
        print(paste("Dim wide fire:", dim(wide_fire)))

        start_index = which(1==names(wide_fire))-1
        
        for (x in 1:max(grid_index)) {
            # pull list of NAs and fire identifiers for each col/grid cell
            full_lst <- pull(wide_fire, x+start_index)
            # list of indexes for present fires
            fire_idx <- which(!is.na(full_lst))
            # list of present fire identifiers
            lst <- full_lst[!is.na(full_lst)]
            n_fires <- length(lst)
            # if no fires, NA or 0 as applicable
            if (length(lst) == 0) {
                fire_lst_tib_mp$fire_lstmp[x] = NA
                fire_lst_tib_mp$max_acresmp[x] = 0
                fire_lst_tib_mp$med_acresmp[x] = 0
                fire_lst_tib_mp$min_mgmtmp[x] = NA
                fire_lst_tib_mp$max_durationmp[x] = 0
                fire_lst_tib_mp$med_durationmp[x] = 0
            # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
            } else {
                fire_lst_tib_mp$fire_lstmp[x] = paste(unlist(lst), collapse=", ")
                fire_lst_tib_mp$max_acresmp[x] = max(wide_fire$acres[fire_idx], na.rm=TRUE)
                fire_lst_tib_mp$med_acresmp[x] = median(wide_fire$acres[fire_idx], na.rm=TRUE)
                fire_lst_tib_mp$min_mgmtmp[x] = min(wide_fire$fire_management_n[fire_idx], na.rm=TRUE)
                fire_lst_tib_mp$max_durationmp[x] = max(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
                fire_lst_tib_mp$med_durationmp[x] = median(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
            }
            fire_lst_tib_mp$n_firesmp[x] = n_fires
        }

    } else {
        # if no fires, populate cells with all 0/NA
        print("No fires recorded this year.")
        for (x in 1:max(grid_index)) {
            fire_lst_tib_mp$fire_lstmp[x] = NA
            fire_lst_tib_mp$max_acresmp[x] = 0
            fire_lst_tib_mp$med_acresmp[x] = 0
            fire_lst_tib_mp$min_mgmtmp[x] = NA
            fire_lst_tib_mp$max_durationmp[x] = 0
            fire_lst_tib_mp$med_durationmp[x] = 0
            fire_lst_tib_mp$n_firesmp[x] = 0
        }
        
    }

    # add current year to full tibble
    full_tib_mp <- rbind(full_tib_mp, fire_lst_tib_mp)
}

head(full_tib_mp)

unique(full_tib_mp$year)

full_tib_mp %>%
    group_by(year) %>%
    summarize(
        n_fire = sum(n_firesmp),
        avg_med_acres = mean(med_acresmp),
        avg_max_acres = mean(max_acresmp)
    )

```

### Point wildfires

```{r}
# initialize empty tibble to hold all values split by year
full_tib_pt <- tibble(
    row_id = numeric(),
    year = numeric(),
    index = numeric()
)

for (i in START_YEAR:END_YEAR) {
    print(paste("Year:", i))
    # initialize empty tibble for current year
    fire_lst_tib_pt <- tibble(grid_year) %>% select(row_id, year, index) %>% filter(year==i)
    # intersecting fires by grid for current year
    fire_pt_int <- st_contains(grid_base, fire_pt %>% filter(Clndr_Y==i), sparse=FALSE)
    # convert to tibble
    fire_pt_tib <- cbind(tibble(data.frame(fire_pt_int)), grid_index)

    # fire_pt with only fires from current year
    fire_pt_curr <- fire_pt %>% filter(Clndr_Y == i)

    if (nrow(fire_pt_curr) > 0) {

        print(paste("N fires:", nrow(fire_pt_curr)))

        # long fire with information from every fire in current year
        long_fire <- fire_pt_tib %>%
            pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire")
            
        if (nrow(fire_pt_curr) > 1) {
            long_fire <- long_fire %>%
                mutate(fire_index = as.numeric(str_remove(fire_index, "X")))
        } else {
           long_fire <- long_fire %>%
                mutate(fire_index = 1)
        }

        long_fire <- long_fire %>%
            mutate(
                unique_id = fire_pt_curr$UNQE_FI[fire_index],
                acres = fire_pt_curr$Clclt_A[fire_index],
                acres_tribal = fire_pt_curr$AcrsTrb[fire_index],
                fire_management_n = fire_pt_curr$fr_mgm_[fire_index],
                duration_hrs = fire_pt_curr$drtn_hr[fire_index]
        )

        print(paste("Long fire dim:", dim(long_fire)))

        # wide fire that only includes uid if fire is present in cell
        wide_fire <- long_fire %>%
            mutate(
                limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
            ) %>%
            pivot_wider(names_from = grid_index, values_from = limited_uid)

        print(paste("Wide fire dim:", dim(wide_fire)))

        start_index = which(1==names(wide_fire))-1
        
        for (x in 1:max(grid_index)) {
            # pull list of NAs and fire identifiers for each col/grid cell
            full_lst <- pull(wide_fire, x+start_index)
            # list of indexes for present fires
            fire_idx <- which(!is.na(full_lst))
            # list of present fire identifiers
            lst <- full_lst[!is.na(full_lst)]
            n_fires <- length(lst)
            # if no fires, NA or 0 as applicable
            if (length(lst) == 0) {
                fire_lst_tib_pt$fire_lstpt[x] = NA
                fire_lst_tib_pt$max_acrespt[x] = 0
                fire_lst_tib_pt$med_acrespt[x] = 0
                fire_lst_tib_pt$max_acres_tribal[x] = 0
                fire_lst_tib_pt$med_acres_tribal[x] = 0
                fire_lst_tib_pt$min_mgmtpt[x] = NA
                fire_lst_tib_pt$max_durationpt[x] = 0
                fire_lst_tib_pt$med_durationpt[x] = 0
            # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
            } else {
                fire_lst_tib_pt$fire_lstpt[x] = paste(unlist(lst), collapse=", ")
                fire_lst_tib_pt$max_acrespt[x] = max(wide_fire$acres[fire_idx], na.rm=TRUE)
                fire_lst_tib_pt$med_acrespt[x] = median(wide_fire$acres[fire_idx], na.rm=TRUE)
                fire_lst_tib_pt$max_acres_tribal[x] = max(wide_fire$acres_tribal[fire_idx], na.rm=TRUE)
                fire_lst_tib_pt$med_acres_tribal[x] = median(wide_fire$acres_tribal[fire_idx], na.rm=TRUE)
                fire_lst_tib_pt$min_mgmtpt[x] = min(wide_fire$fire_management_n[fire_idx], na.rm=TRUE)
                fire_lst_tib_pt$max_durationpt[x] = max(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
                fire_lst_tib_pt$med_durationpt[x] = median(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
            }
            fire_lst_tib_pt$n_firespt[x] = n_fires
        }
    } else {
        print("No fires recorded this year.")
        for (x in 1:max(grid_index)) {
            fire_lst_tib_pt$fire_lstpt[x] = NA
            fire_lst_tib_pt$max_acrespt[x] = 0
            fire_lst_tib_pt$med_acrespt[x] = 0
            fire_lst_tib_pt$max_acres_tribal[x] = 0
            fire_lst_tib_pt$med_acres_tribal[x] = 0
            fire_lst_tib_pt$min_mgmtpt[x] = NA
            fire_lst_tib_pt$max_durationpt[x] = 0
            fire_lst_tib_pt$med_durationpt[x] = 0
            fire_lst_tib_pt$n_firespt[x] = 0
        }
    }
    # add current year to full tibble
    full_tib_pt <- rbind(full_tib_pt, fire_lst_tib_pt)
}

head(full_tib_pt)

unique(full_tib_pt$year)

full_tib_pt %>%
    arrange(desc(max_acrespt))

full_tib_pt %>%
    group_by(year) %>%
    summarize(
        n_fire = sum(n_firespt),
        avg_med_acres = mean(med_acrespt, na.rm=TRUE),
        avg_max_acres = mean(max_acrespt, na.rm=TRUE)
    )

```

## Prescribed Burns

### Multipolygon RX

```{r}
# initialize empty tibble to hold all values split by year
full_tib_rxmp <- tibble(
    row_id = numeric(),
    year = numeric(),
    index = numeric()
)

# note: this takes a couple of minutes to run
# this is not a surprise as it's testing however many fires against 10k cells 10 times. 
# there may be a more efficient way to do this but idk what it is

for (i in START_YEAR:END_YEAR) {
    print(paste("Year:", i))
    # initialize empty tibble for current year
    rx_lst_tib_mp <- tibble(grid_year) %>% select(row_id, year, index) %>% filter(year==i)
    # intersecting fires by grid for current year
    rx_mp_int <- st_intersects(grid_base, rx_fire_mp %>% filter(Clndr_Y==i), sparse=FALSE)
    # convert to tibble
    rx_mp_tib <- cbind(tibble(data.frame(rx_mp_int)), grid_index)

    # fire_mp with only fires from current year
    rx_mp_curr <- rx_fire_mp %>% filter(Clndr_Y == i)

    if (nrow(rx_mp_curr) > 0) {
        print(paste("N rx fires:", dim(rx_mp_curr)[1]))
        # long fire with information from every fire in current year
        long_fire <- rx_mp_tib %>%
            pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") 
            
        if (nrow(rx_mp_curr) > 1) {
            long_fire <- long_fire %>%
                mutate(fire_index = as.numeric(str_remove(fire_index, "X")))
        } else {
            long_fire <- long_fire %>%
                mutate(fire_index = 1)
        }

        long_fire <- long_fire %>%
            mutate(
                unique_id = rx_mp_curr$UNQE_FI[fire_index],
                acres = rx_mp_curr$GIS_ACR[fire_index],
                fire_management_n = rx_mp_curr$fr_mgm_[fire_index],
                duration_hrs = rx_mp_curr$drtn_hr[fire_index]
            )

        # wide fire that only includes uid if fire is present in cell
        wide_fire <- long_fire %>%
            mutate(
                limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
            ) %>%
            pivot_wider(names_from = grid_index, values_from = limited_uid)

        start_index = which(1==names(wide_fire))-1
        
        for (x in 1:max(grid_index)) {
            # pull list of NAs and fire identifiers for each col/grid cell
            full_lst <- pull(wide_fire, x+start_index)
            # list of indexes for present fires
            fire_idx <- which(!is.na(full_lst))
            # list of present fire identifiers
            lst <- full_lst[!is.na(full_lst)]
            n_fires <- length(lst)

            # if no fires, NA or 0 as applicable
            if (length(lst) == 0) {
                rx_lst_tib_mp$rx_lstmp[x] = NA
                rx_lst_tib_mp$max_acresmp_rx[x] = 0
                rx_lst_tib_mp$med_acresmp_rx[x] = 0
                rx_lst_tib_mp$min_mgmtmp_rx[x] = NA
                rx_lst_tib_mp$max_durationmp_rx[x] = 0
                rx_lst_tib_mp$med_durationmp_rx[x] = 0
            # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
            } else {
                rx_lst_tib_mp$rx_lstmp[x] = paste(unlist(lst), collapse=", ")
                rx_lst_tib_mp$max_acresmp_rx[x] = max(wide_fire$acres[fire_idx], na.rm=TRUE)
                rx_lst_tib_mp$med_acresmp_rx[x] = median(wide_fire$acres[fire_idx], na.rm=TRUE)
                rx_lst_tib_mp$min_mgmtmp_rx[x] = min(wide_fire$fire_management_n[fire_idx], na.rm=TRUE)
                rx_lst_tib_mp$max_durationmp_rx[x] = max(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
                rx_lst_tib_mp$med_durationmp_rx[x] = median(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
            }
            rx_lst_tib_mp$n_rxmp[x] = n_fires
        }

    } else {
        print("No prescribed burns recorded this year.")
        for (x in 1:max(grid_index)) {
            rx_lst_tib_mp$rx_lstmp[x] = NA
            rx_lst_tib_mp$max_acresmp_rx[x] = 0
            rx_lst_tib_mp$med_acresmp_rx[x] = 0
            rx_lst_tib_mp$min_mgmtmp_rx[x] = NA
            rx_lst_tib_mp$max_durationmp_rx[x] = 0
            rx_lst_tib_mp$med_durationmp_rx[x] = 0
            rx_lst_tib_mp$n_rxmp[x] = 0
        }
    
    }

    # add current year to full tibble
    full_tib_rxmp <- rbind(full_tib_rxmp, rx_lst_tib_mp)

}

head(full_tib_rxmp)

unique(full_tib_rxmp$year)

full_tib_rxmp %>%
    group_by(year) %>%
    summarize(
        n_rx = sum(n_rxmp),
        avg_med_acres = mean(med_acresmp_rx),
        avg_max_acres = mean(max_acresmp_rx)
    )

```

### Point RX

```{r}
# initialize empty tibble to hold all values split by year
full_tib_rxpt <- tibble(
    row_id = numeric(),
    year = numeric(),
    index = numeric()
)

for (i in START_YEAR:END_YEAR) {
    print(paste("Year:", i))
    # initialize empty tibble for current year
    rx_lst_tib_pt <- tibble(grid_year) %>% select(row_id, year, index) %>% filter(year==i)
    # intersecting fires by grid for current year
    rx_pt_int <- st_contains(grid_base, rx_fire_pt %>% filter(Clndr_Y==i), sparse=FALSE)
    # convert to tibble
    rx_pt_tib <- cbind(tibble(data.frame(rx_pt_int)), grid_index)

    # fire_pt with only fires from current year
    rx_pt_curr <- rx_fire_pt %>% filter(Clndr_Y == i)

    if (nrow(rx_pt_curr) > 0) {
        print(paste("N rx fires:", dim(rx_pt_curr)[1]))
        # long fire with information from every fire in current year
        long_fire <- rx_pt_tib %>%
            pivot_longer(!grid_index, names_to="fire_index", values_to="is_fire") 
            
        if (nrow(rx_pt_curr) > 1) {
            long_fire <- long_fire %>%
                mutate(fire_index = as.numeric(str_remove(fire_index, "X")))
        } else {
            long_fire <- long_fire %>%
                mutate(fire_index = 1)
        }

        long_fire <- long_fire %>%
            mutate(
                unique_id = rx_pt_curr$UNQE_FI[fire_index],
                acres = rx_pt_curr$Clclt_A[fire_index],
                acres_tribal = rx_pt_curr$AcrsTrb[fire_index],
                fire_management_n = rx_pt_curr$fr_mgm_[fire_index],
                duration_hrs = rx_pt_curr$drtn_hr[fire_index]
            )

        # wide fire that only includes uid if fire is present in cell
        wide_fire <- long_fire %>%
            mutate(
                limited_uid = ifelse(is_fire==TRUE, unique_id, NA)
            ) %>%
            pivot_wider(names_from = grid_index, values_from = limited_uid)

        print(dim(wide_fire)) # twice as many rows as there are fires
        # this is because each has is_fire=FALSE and TRUE

        start_index = which(1==names(wide_fire))-1
        
        for (x in 1:max(grid_index)) {
            # pull list of NAs and fire identifiers for each col/grid cell
            full_lst <- pull(wide_fire, x+start_index)
            # list of indexes for present fires
            fire_idx <- which(!is.na(full_lst))
            # list of present fire identifiers
            lst <- full_lst[!is.na(full_lst)]
            n_fires <- length(lst)
            # if no fires, NA or 0 as applicable
            if (length(lst) == 0) {
                rx_lst_tib_pt$rx_lstpt[x] = NA
                rx_lst_tib_pt$max_acrespt_rx[x] = 0
                rx_lst_tib_pt$med_acrespt_rx[x] = 0
                rx_lst_tib_pt$max_acres_tribal_rx[x] = 0
                rx_lst_tib_pt$med_acres_tribal_rx[x] = 0
                rx_lst_tib_pt$min_mgmtpt_rx[x] = NA
                rx_lst_tib_pt$max_durationpt_rx[x] = 0
                rx_lst_tib_pt$med_durationpt_rx[x] = 0
            # if fires present, add list of identifiers, max acres, min mgmt complexity, max duration
            } else {
                rx_lst_tib_pt$rx_lstpt[x] = paste(unlist(lst), collapse=", ")
                rx_lst_tib_pt$max_acrespt_rx[x] = max(wide_fire$acres[fire_idx], na.rm=TRUE)
                rx_lst_tib_pt$med_acrespt_rx[x] = median(wide_fire$acres[fire_idx], na.rm=TRUE)
                rx_lst_tib_pt$max_acres_tribal_rx[x] = max(wide_fire$acres_tribal[fire_idx], na.rm=TRUE)
                rx_lst_tib_pt$med_acres_tribal_rx[x] = median(wide_fire$acres_tribal[fire_idx], na.rm=TRUE)
                rx_lst_tib_pt$min_mgmtpt_rx[x] = min(wide_fire$fire_management_n[fire_idx], na.rm=TRUE)
                rx_lst_tib_pt$max_durationpt_rx[x] = max(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
                rx_lst_tib_pt$med_durationpt_rx[x] = median(wide_fire$duration_hrs[fire_idx], na.rm=TRUE)
            }
            rx_lst_tib_pt$n_rxpt[x] = n_fires
        }
    } else {
        print("No prescribed burns recorded this year.")
        for (x in 1:max(grid_index)) {
            rx_lst_tib_pt$rx_lstpt[x] = NA
            rx_lst_tib_pt$max_acrespt_rx[x] = 0
            rx_lst_tib_pt$med_acrespt_rx[x] = 0
            rx_lst_tib_pt$max_acres_tribal_rx[x] = 0
            rx_lst_tib_pt$med_acres_tribal_rx[x] = 0
            rx_lst_tib_pt$min_mgmtpt_rx[x] = NA
            rx_lst_tib_pt$max_durationpt_rx[x] = 0
            rx_lst_tib_pt$med_durationpt_rx[x] = 0
            rx_lst_tib_pt$n_rxpt[x] = 0
        }
    }
    # add current year to full tibble
    full_tib_rxpt <- rbind(full_tib_rxpt, rx_lst_tib_pt)
}

head(full_tib_rxpt)

unique(full_tib_rxpt$year)

full_tib_rxpt %>%
    group_by(year) %>%
    summarize(
        n_rx = sum(n_rxpt),
        avg_med_acres = mean(med_acrespt_rx, na.rm=TRUE),
        avg_max_acres = mean(max_acrespt_rx, na.rm=TRUE)
    )

```

## Combining All Tibbles

```{r}
# keeping year and index initially for validation
# removing here for ease of cbind
full_tib_mp <- full_tib_mp %>% select(-c(year, index))
full_tib_pt <- full_tib_pt %>% select(-c(year, index))
full_tib_rxmp <- full_tib_rxmp %>% select(-c(year, index))
full_tib_rxpt <- full_tib_rxpt %>% select(-c(year, index))

full_tib <- full_join(full_tib_mp, full_tib_pt, by="row_id")
full_tib <- full_join(full_tib, full_tib_rxmp, by="row_id")
full_tib <- full_join(full_tib, full_tib_rxpt, by="row_id")

```

```{r}
names(full_tib)
full_tib <- full_tib %>%
    rowwise() %>%
    mutate(
        max_acres = max(max_acresmp, max_acrespt, na.rm=TRUE),
        max_duration = max(max_durationmp, max_durationpt, na.rm=TRUE),
        min_mgmt = min(min_mgmtmp, min_mgmtpt, na.rm=TRUE),
        max_acres_rx = max(max_acresmp_rx, max_acrespt_rx, na.rm=TRUE),
        max_duration_rx = max(max_durationmp_rx, max_durationpt_rx, na.rm=TRUE),
        min_mgmt_rx = min(min_mgmtmp_rx, min_mgmtpt_rx, na.rm=TRUE)
    )

full_tib <- full_tib %>%
    mutate(
        min_mgmt = na_if(min_mgmt, Inf),
        min_mgmtmp = na_if(min_mgmtmp, Inf),
        min_mgmtpt = na_if(min_mgmtpt, Inf),
        min_mgmt_rx = na_if(min_mgmt_rx, Inf),
        min_mgmtmp_rx = na_if(min_mgmtmp_rx, Inf),
        min_mgmtpt_rx = na_if(min_mgmtpt_rx, Inf)
    )

```

```{r}
time_grid <- full_join(grid_year, full_tib, by="row_id")

names(time_grid)
dim(time_grid)
```

```{r}
object.size(time_grid)
st_write(time_grid, "../DATA-510/time_grid.shp", append=FALSE)

time_grid_tibble <- tibble(time_grid %>% select(-geometry))
object.size(time_grid_tibble)
write_csv(time_grid_tibble, "../DATA-510/time_grid.csv")
write_csv(time_grid_tibble, "data/processed_grid/time_grid.csv")
```

Unfortunately the time grid shapefile is too large to upload to github! This is very sad. We will have to find another solution!